---
title: "Combined_Fig6_ImmunePhenotypePrimaryTumor_DA"
author: "LauraKuett"
date: "2021-02-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
## Introduction

This script check for differences between inflamed and excluded primary tumors

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and data

First, we will load the needed libraries and the `SingleCellExperiment` object, that was created in the previous section.

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(dittoSeq)
library(scales)
library(tidyverse)
library(ggpubr)
library(cowplot)
library(scran)
library(CATALYST)
library(edgeR)
library(compositions)
library(vegan)
library(ComplexHeatmap)

```

### Load local functions(modified from the CATALYST package)

```{r}
local_split_cells <- function(x, by) {
    stopifnot(is.character(by), by %in% colnames(colData(x)))
    cd <- as.data.frame(colData(x))
    dt <- data.table::data.table(cd, i = seq_len(ncol(x)))
    dt_split <- split(dt, by = by, sorted = TRUE, flatten = FALSE)
    purrr::map_depth(dt_split, length(by), "i")
}

local_agg <- function(x, 
    by = c("cluster_id", "sample_id"), 
    fun = c("median", "mean", "sum"),
    assay = "exprs") {
    fun <- match.arg(fun)
    y <- assay(x, assay)
    if (fun == "median" && !is.matrix(y))
        y <- as.matrix(y)
    fun <- switch(fun, 
        median = rowMedians, 
        mean = rowMeans, 
        sum = rowSums)
    cs <- local_split_cells(x, by)
    pb <- map_depth(cs, -1, function(i) {
        if (length(i) == 0) return(numeric(nrow(x)))
        fun(y[, i, drop = FALSE])
    })
    purrr::map_depth(pb, -2, function(u) as.matrix(data.frame(
        u, row.names = rownames(x), check.names = FALSE)))
}


```


### Load T cell panel

```{r}
sce <- readRDS("output/SCEs/Tcell/TcellSCE_subclustered.rds")

sce$patient.tissue <- paste(sce$Patient.ID, sce$Tissue.Type, sep = "_")
sce$patient.tissue.loc <- paste(sce$patient.tissue, sce$Location, sep = "_")

# Read in ROI metadata
ROI_meta <- read.csv("data/immune_images_patient_metadata.csv")
ROI_meta$patient.tissue <- paste(ROI_meta$Patient.ID, ROI_meta$Tissue.Type, sep = "_")
sample_unique <- ROI_meta[!duplicated(ROI_meta$patient.tissue),]
rownames(sample_unique) <- sample_unique$patient.tissue

ROI_meta$patient.tissue.loc <- paste(ROI_meta$patient.tissue, ROI_meta$Location, sep = "_")
ROI_unique <- ROI_meta[!duplicated(ROI_meta$patient.tissue.loc),]
rownames(ROI_unique) <- ROI_unique$patient.tissue.loc

#set up factors for CATALYST PACKAGE
sce$sample_type[sce$Tissue.Type == 'PT'] <- "PT"
sce$sample_type[sce$Tissue.Type == 'Liver'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Bone'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Brain'] <- "M"
sce$sample_type[sce$Tissue.Type == 'SoftTissue'] <- "M"
sce$sample_type <- factor(sce$sample_type)

# Set up necessary fields for CATALYST plotting functions
sce$sample_id <- factor(paste(sce$Patient.ID, sce$sample_type))
sce$Tissue.Type <- factor(sce$Tissue.Type)
sce$Site.of.metastasis <- factor(sce$Site.of.metastasis)
sce$Location <- factor(sce$Location)
sce$Tcell_subtype <- as.factor(sce$Tcell_subtype)
sce$Tcell_cluster <- factor(sce$Tcell_cluster)

#add immune phenotyping information
immune_pheno <- read.csv("data/patient_metadata_encoded.csv")
immune_pheno_name <- immune_pheno$immunePT_PT[match(sce$Patient.ID, immune_pheno$Patient.ID)]
sce$immunePT_PT <- factor(immune_pheno_name)


immune_pheno$immunePT_Met[immune_pheno$immunePT_Met=="x_desert"] <- "desert"
immune_pheno_name1 <- immune_pheno$immunePT_Met[match(sce$Patient.ID, immune_pheno$Patient.ID)]
sce$immunePT_Met <- factor(immune_pheno_name1)

sce_initial <- sce
# Subset immune cells
sce <- sce_initial[,which(sce_initial$cell_class == "immune")]
sce$celltype <- ifelse(sce$celltype == "T_cell", sce$clustername, sce$celltype)


```


### Display patient tissue immune phenotypes in PT and Met

```{r alluvial-plot-tissue-immune-phenotype}
library(ggalluvial)

# Exclude the one patient whose metastatic sites have different annotations
patho1 <- immune_pheno[immune_pheno$Patient.ID != "ID17",]

patho1$links <- paste(patho1$immunePT_PT, patho1$immunePT_Met, sep = "_")
links <- as.data.frame(table(patho1$links))
links <- separate(links, 1, into = c("PT", "Met"), sep = "_" )
links$subject <- c(1:5)

patho2 <- pivot_longer(links, 1:2, names_to = "tissue", values_to = "type")
patho2$tissue <- factor(patho2$tissue, level = c("PT", "Met"))

ggplot(patho2,
       aes(x = tissue, stratum = type, alluvium = subject, y = Freq, fill = type, label = type)) +
  scale_x_discrete(expand = c(.1, .1)) +
  scale_fill_manual(values = c("grey", "orange", "darkred"))+
  geom_flow(alpha = .3) +
  geom_stratum(alpha = .5) +
  geom_text(stat = "stratum", size = 3) +
  theme(legend.position = "none", panel.background=element_blank())



```


### Barplot to show which metastatic sites have which T cell infiltration status
```{r bar-plot-immune-tissue-phenotype}
patho_long <- pivot_longer(patho1, 15:16, names_to = "MetPT", values_to = "ImPh")
patho_long$MetPT <- factor(patho_long$MetPT, levels =c("immunePT_PT","immunePT_Met"))
colors <- sce@metadata$colour_vectors$tissuetype
names(colors)[1] <- "Soft Tissue"
colors["Multiple"] <- "darkgreen"
colors <- colors[colors != "#4C9900"]

ggplot(patho_long, aes(ImPh, fill = Site.of.metastasis))+
  geom_bar(position = "stack")+
  scale_fill_manual(values = colors)+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))+
  facet_wrap(~MetPT)
```


### Check DA for general immune cell subtypes
```{r inflamed-excluded-DA-immune-cells-intratumoral}

sce$cluster_id <- factor(sce$celltype)
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))
sce_filtered1 <- filterSCE(sce, sample_type =="PT")
sce_filtered <- filterSCE(sce_filtered1, Location =="intratumoral")
i = "Excluded "

# Prepare abundance object
abundances <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_filtered)[match(colnames(abundances), sce_filtered$Patient.ID),]
rownames(meta) <- meta$Patient.ID
immunePT_PT <- factor(meta$immunePT_PT)
meta$immunePT_PT <- relevel(x=immunePT_PT, ref = "excluded")
y.ab <- DGEList(abundances, samples=meta,remove.zeros = TRUE,group = meta$immunePT_PT)
# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add tissue type as an additive factor
design <- model.matrix(~immunePT_PT, data = y.ab$samples)

# Calculate Normalization Factors to adjust for composition effect (this assumes that most cell type populations stay constant, which is not necessarily true)
y.ab <- calcNormFactors(y.ab)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=ncol(fit.ab))
summary(decideTests(res))
DA <- topTags(res, n = nrow(abundances))$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs inflamed"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```


# Test if acquired regions mirror the slide-wide immune phenotype classification

### Prepare spatial object relationship data

```{r object-relationships}

# Read object relationship file (output from Cellprofiler)
dat_relation <- data.table::fread("data/immune_Object relationships.csv")
dat_relation$cellA <- paste0(dat_relation$`First Image Number`, "_", dat_relation$`First Object Number`)
dat_relation$cellB <- paste0(dat_relation$`Second Image Number`, "_", dat_relation$`Second Object Number`)
```

```{r ROI-immune-phenotypes-withoutMargin, fig.width=4, fig.height=5}

patho_short <- patho1[,c(1,15,16)]
colnames(patho_short) <- c("Patient.ID", "PT", "M")
patho_short <- pivot_longer(patho_short, 2:3, names_to = "site", values_to = "imPh")
patho_short$patient.site <- paste(patho_short$Patient.ID, patho_short$site, sep = " ")

# Exclude patients not present in the pathology list
sce_sub <- sce_initial[,which(sce_initial$Patient.ID %in% patho1$Patient.ID)]

sce_sub$imPh <- patho_short$imPh[match(sce_sub$sample_id, patho_short$patient.site)]

#exclude margin images
sce_nomarg <- sce_sub[,which(sce_sub$Location == "intratumoral")]

cur_sce <- data.frame(colData(sce_nomarg))
df <- data.frame(matrix(nrow=0, ncol =4))

for (i in unique(cur_sce$sample_id)){
  cur_sce_i <- cur_sce[cur_sce[,"sample_id"] == i,]
  
  # get T cell and tumor cell IDs
  cellIDs_Tcell <- cur_sce_i[cur_sce_i[,"clustername"] == "T_CD8",]
  Tcell_shortID <- paste0(cellIDs_Tcell$ImageNumber, "_", cellIDs_Tcell$CellNumber)
  cellIDs_tumor <- cur_sce_i[cur_sce_i[,"celltype"] == "epithelial",]
  tumor_shortID <- paste0(cellIDs_tumor$ImageNumber, "_", cellIDs_tumor$CellNumber)
  
  # subset neighbourhood list to include only `T cell <-> tumor cell` interactions
  dat_relation_sub <- dat_relation %>% filter((cellA %in% Tcell_shortID & cellB %in% tumor_shortID))
  # count interactions
  intnr_i <- nrow(dat_relation_sub)
  
  # divide through T cell nr
  intnr_Tcell_prop <- intnr_i/length(Tcell_shortID)
  # divide through T cell nr
  intnr_tumor_prop <- intnr_i/length(tumor_shortID)
  
  new_row <- c(i, intnr_i, intnr_Tcell_prop, intnr_tumor_prop)
  df <- rbind(df, new_row)
} 

colnames(df) <- c("patient.site", "total_interactions", "Tcell_prop", "tumor_prop")
df[,2:4] <- df[,2:4] %>% mutate_if(is.character,as.numeric)

df_merged <- merge(df, patho_short, by = "patient.site")



### Direct T cell interactions per tumor cell
my_comparisons <- list( c("inflamed", "excluded"), c("inflamed", "desert"), c("desert", "excluded"))

ggplot(df_merged, aes(imPh, tumor_prop))+
    geom_boxplot()+
    geom_point(position=position_dodge(width=0.75))+
    stat_compare_means(method = "wilcox.test", aes(label=..p.format..), comparisons = my_comparisons)+
    theme(axis.line.x = element_line(colour = "black", size = 0.25),
        axis.line.y = element_line(colour = "black", size = 0.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank(),
        strip.background = element_blank()) + ggtitle("Average number of direct CD8+ T cell \ninteractions per tumor cell")


```




## Check DA between inflamed and exhausted phenotype for T cell metaclusters in intratumoral region

```{r inflamed-excluded-DA-intratumoral-Tcells}

sce$cluster_id <- sce$Tcell_subtype
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))
sce_filtered1 <- filterSCE(sce, sample_type =="PT")
sce_filtered2 <- filterSCE(sce_filtered1, cluster_id !="none")
sce_filtered <- filterSCE(sce_filtered2, Location =="intratumoral")

i = "Excluded "

# Prepare abundance object
abundances <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_filtered)[match(colnames(abundances), sce_filtered$Patient.ID),]
rownames(meta) <- meta$Patient.ID
immunePT_PT <- factor(meta$immunePT_PT)
meta$immunePT_PT <- relevel(x=immunePT_PT, ref = "excluded")
y.ab <- DGEList(abundances, samples=meta,remove.zeros = TRUE,group = meta$immunePT_PT)
# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add tissue type as an additive factor
design <- model.matrix(~immunePT_PT, data = y.ab$samples)
# Calculate Normalization Factors to adjust for composition effect (this assumes that most cell type populations stay constant, which is not necessarily true)
y.ab <- calcNormFactors(y.ab)


# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=ncol(fit.ab))
summary(decideTests(res))
DA <- topTags(res, n = nrow(abundances))$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs inflamed"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))

```


### Load macrophage panel
```{r}

sce_i <- readRDS("output/SCEs/myeloid/myeloidSCE_subclustered.rds")

#Remove patient 80 due to too few cells
sce_i <- sce_i[,sce_i$Patient.ID != "ID80"]

sce2 <- filterSCE(sce_i,cell_class=="immune")
sce2$cluster_id <- factor(sce2$M_cluster)

markers_to_analyse<- c("CD38", "HLA-DR", "CLEC9A", "CD68", "CD163", "CD11b", "CD204", "GLUT1", "CD11c", "CD1c", "CD16", "MMP9", "CD274", "STING", "IDO", "CD33", "Arginase-1", "CD209", "CD40", "CD4", "CD14", "CD206", "LAMP3", "MPO", "CD303")

sce <- filterSCE(sce2, rownames(sce2) %in% markers_to_analyse)


#set up factors for CATALYST PACKAGE
sce$sample_type[sce$Tissue.Type == 'PT'] <- "PT"
sce$sample_type[sce$Tissue.Type == 'Liver'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Bone'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Brain'] <- "M"
sce$sample_type[sce$Tissue.Type == 'SoftTissue'] <- "M"
sce$sample_type <- factor(sce$sample_type)
# Set up necessary fields for CATALYST plotting functions
sce$sample_id <- factor(paste(sce$Patient.ID, sce$sample_type))


sce$Tissue.Type <- factor(sce$Tissue.Type)
sce$Site.of.metastasis <- factor(sce$Site.of.metastasis)
sce$Location <- factor(sce$Location)
sce$clustername <- as.factor(sce$clustername)
sce$M_cluster <- as.factor(sce$M_cluster)

sce$cluster_id<- as.factor(sce$M_cluster)
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))

#add immune phenotyping information
immune_pheno <- read.csv("data/patient_metadata_encoded.csv")
immune_pheno_name <- immune_pheno$immunePT_PT[match(sce$Patient.ID, immune_pheno$Patient.ID)]
sce$immunePT_PT <- factor(immune_pheno_name)
immune_pheno_name1 <- immune_pheno$immunePT_Met[match(sce$Patient.ID, immune_pheno$Patient.ID)]
sce$immunePT_Met <- factor(immune_pheno_name1)

```

### DA for macrophage subtypes in non-margin region

```{r myeloid-cell-DA-inflamed-excluded-intratumoral}
sce_filtered1 <- filterSCE(sce, sample_type =="PT")
sce_filtered2 <- filterSCE(sce_filtered1, Location !="margin")
sce_filtered <- filterSCE(sce_filtered2, cluster_id !="none")


i = "Excluded "

# Prepare abundance object
abundances <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_filtered)[match(colnames(abundances), sce_filtered$Patient.ID),]
rownames(meta) <- meta$Patient.ID
immunePT_PT <- factor(meta$immunePT_PT)
meta$immunePT_PT <- relevel(x=immunePT_PT, ref = "excluded")
y.ab <- DGEList(abundances, samples=meta,remove.zeros = TRUE,group = meta$immunePT_PT)
# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add tissue type as an additive factor
design <- model.matrix(~immunePT_PT, data = y.ab$samples)

# Calculate Normalization Factors to adjust for composition effect (this assumes that most cell type populations stay constant, which is not necessarily true)
y.ab <- calcNormFactors(y.ab)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=ncol(fit.ab))
summary(decideTests(res))
DA <- topTags(res, n = nrow(abundances))$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs inflamed"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

### Compare epithelial cell expression
```{r inflamed-excluded-DA-intratumoral-tumor-cells}

sce_epithelial <- readRDS("output/SCEs/tumor/tumorSCE_epithelial_clustered.rds")

markers_to_analyse<- c("HLA-ABC","HH3", "SMA","panCK","Vimentin","CK8_18_19",
                     "ER", "CK14","p53", "GLUT1","Bcl-2" ,"HER2", "CK5" ,"CD274", "AR","GATA3","CK7", 
                     "CAIX" ,"Fibronectin", "Ki-67_Er168","EGFR","p-S6",
                     "mTOR","E/P-Cadherin","p-Rb","cleaved_CP")

sce4 <- filterSCE(sce_epithelial, rownames(sce_epithelial) %in% markers_to_analyse)

counts_per_sample <- table(sce_epithelial$sample_type, sce_epithelial$Patient.ID)
samples_with_few_cells <- colSums(counts_per_sample<50)
patient_to_exclude <- rownames(data.frame(samples_with_few_cells[samples_with_few_cells==1]))

sce_tumor <- filterSCE(sce4,!(sce4$Patient.ID %in% patient_to_exclude))

#add immune phenotyping information
immune_pheno <- read.csv("data/patient_metadata_encoded.csv")
immune_pheno_name <- immune_pheno$immunePT_PT[match(sce_tumor$Patient.ID, immune_pheno$Patient.ID)]
sce_tumor$immunePT_PT <- factor(immune_pheno_name)
immune_pheno_name1 <- immune_pheno$immunePT_Met[match(sce_tumor$Patient.ID, immune_pheno$Patient.ID)]
sce_tumor$immunePT_Met <- factor(immune_pheno_name1)

#Repeat tissue type as condition for plotting functions in CATALYST
sce_tumor$condition <- factor(sce_tumor$immunePT_PT)
sce_tumor$Patient.ID <- factor(sce_tumor$Patient.ID)
sce_tumor$Location <- factor(sce_tumor$Location)

sce_filtered <- filterSCE(sce_tumor, sample_type =="PT")

i = "Excluded "

# Prepare abundance object
abundances <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_filtered)[match(colnames(abundances), sce_filtered$Patient.ID),]
rownames(meta) <- meta$Patient.ID
immunePT_PT <- factor(meta$immunePT_PT)
meta$immunePT_PT <- relevel(x=immunePT_PT, ref = "excluded")
y.ab <- DGEList(abundances, samples=meta,remove.zeros = TRUE,group = meta$immunePT_PT)
# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add tissue type as an additive factor
design <- model.matrix(~immunePT_PT, data = y.ab$samples)

# Calculate Normalization Factors to adjust for composition effect (this assumes that most cell type populations stay constant, which is not necessarily true)
y.ab <- calcNormFactors(y.ab)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=ncol(fit.ab))
summary(decideTests(res))
DA <- topTags(res, n = nrow(abundances))$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs inflamed"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))

```

