---
title: "ImmuneCombined_04_dissimilarity_sameMet"
author: "SandraTietscher"
date: "2022-01-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

This script calculates dissimilarity scores for randomly paired metastatic samples from the same metastatic site, and compares them to dissimilarity scores for matched PT & met samples.

Not entirely sure if this approach is valid, because metastatic samples have generally far fewer cells, and because there is only a small number of possible Met&Met pairs to permute over.

# Load libraries and data
```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(dittoSeq)
library(scales)
library(tidyverse)
library(ggpubr)
library(cowplot)
library(scran)
library(CATALYST)
library(edgeR)
library(vegan)
library(ComplexHeatmap)


# Read immune sce
sce_immune <- readRDS("output/SCEs/Tcell/TcellSCE_subclustered.rds")
sce_immune$patient.tissue <- paste(sce_immune$Patient.ID, sce_immune$Tissue.Type, sep = "_")
sce_immune$MetPT <- ifelse(sce_immune$Tissue.Type == "PT", "PT", "Met")
sce_immune$patient.MetPT <- paste(sce_immune$Patient.ID, sce_immune$MetPT, sep = "_")

# Read macrophage sce
sce_M <- readRDS("output/SCEs/myeloid/myeloidSCE_M.rds")
sce_M$patient.tissue <- paste(sce_M$Patient.ID, sce_M$Tissue.Type, sep = "_")
sce_M$MetPT <- ifelse(sce_M$Tissue.Type == "PT", "PT", "Met")
sce_M$patient.MetPT <- paste(sce_M$Patient.ID, sce_M$MetPT, sep = "_")

# Read in ROI metadata for immune panel
ROI_meta <- read.csv("data/immune_images_patient_metadata_encoded.csv")
ROI_meta$patient.tissue <- paste(ROI_meta$Patient.ID, ROI_meta$Tissue.Type, sep = "_")
sample_unique <- ROI_meta[!duplicated(ROI_meta$patient.tissue),]
rownames(sample_unique) <- sample_unique$patient.tissue


# Read in ROI metadata for myeloid panel
ROI_meta_M <- read.csv("data/macrophage_images_patient_metadata_encoded.csv")

ROI_meta_M$patient.tissue <- paste(ROI_meta_M$Patient.ID, ROI_meta_M$Tissue.Type, sep = "_")
sample_unique_M <- ROI_meta_M[!duplicated(ROI_meta_M$patient.tissue),]
rownames(sample_unique_M) <- sample_unique_M$patient.tissue

```

# Subset: Use only intratumoral samples

```{r subset-intratumoral}
sce_immune_intra <- filterSCE(sce_immune, Location == "intratumoral")
sce_M_intra <- filterSCE(sce_M, Location == "intratumoral")
```

# Subset: Exclude samples with multiple metastatic sites

```{r exclude-multiple}
sce_immune_intra <- filterSCE(sce_immune_intra, Site.of.metastasis != "Multiple")
sce_M_intra <- filterSCE(sce_M_intra, Site.of.metastasis != "Multiple")
```

# General immune cell abundances

### Calculate Aitchison distance between matched samples

For PTs with multiple matched metastatic sites, these sites are pooled before calculating the Aitchison distance between sample pairs.

```{r immune-aitchison}
sce_im <- filterSCE(sce_immune_intra, celltype %in% c("NK_cell", "T_cell", "myeloid", "plasma_cell", "B_cell", "neutrophil"))
sce_met <- filterSCE(sce_im,Tissue.Type!="PT")
sce_PT <- filterSCE(sce_im,Tissue.Type=="PT")

# Calculate centered log-transform of abundances
ncs_met <- table(sce_met$patient.MetPT, sce_met$celltype)
fqs_met  <- log(prop.table(ncs_met + 1, 1), 2)
clr_met  <- as.data.frame.matrix(fqs_met-rowMeans(fqs_met))

ncs_PT <- table(sce_PT$patient.MetPT, sce_PT$celltype)
fqs_PT <- log(prop.table(ncs_PT+ 1, 1), 2)
clr_PT <- as.data.frame.matrix(fqs_PT-rowMeans(fqs_PT))

clr_met$patient.MetPT <- rownames(clr_met)
clr_met <- separate(clr_met, patient.MetPT, into = c("patient", "MetPT"), sep = "_")

clr_PT$patient.MetPT <- rownames(clr_PT)
clr_PT <- separate(clr_PT, patient.MetPT, into = c("patient", "MetPT"), sep = "_")


# Create empty dissimilarity data frame
diss.obs <- as.data.frame(matrix(ncol = 2, nrow = length(rownames(clr_met)), dimnames = list(rownames(clr_met), c("patient.MetPT", "diss_to_PT"))))

# Loop over metastatic tissues and compute dissimilarity to matched PT
for(i in rownames(clr_met)){
 patientID <- clr_met[i,"patient"]
 abundances <- rbind(clr_met[i,], filter(clr_PT, patient == patientID))
 if(nrow(abundances) == 2) {
   diss <- vegdist(abundances[,1:6],method = 'euclidean', binary = FALSE,na.rm = TRUE)
    diss.obs[i,] <- c(i, diss)
 }
}

# Remove NAs (samples for which no PT data is available)
diss.obs <- na.exclude(diss.obs)
diss.obs <- separate(diss.obs, patient.MetPT, into = c("patient", "MetPT"), sep = "_")
diss.obs$site.of.metastasis <- sample_unique$Site.of.metastasis[match(diss.obs$patient, sample_unique$Patient.ID)]
diss.obs$diss_to_PT <- as.numeric(diss.obs$diss_to_PT)
```

### Random matching of Mets to other Mets of the same site

```{r immune-permutations}

# Create empty data frame for 1000 permutations
diss.perm <- as.data.frame(matrix(ncol = 1000, nrow = length(rownames(diss.obs)), dimnames = list(rownames(diss.obs), c(1:1000))))


# Loop for permutations
for (zz in 1:1000) {
  
  # Loop over metastatic tissues and compute dissimilarity to random other metastatis from the same site
  for(i in 1:length(rownames(diss.obs))){
    sample <- rownames(diss.obs)[i]
    site <- diss.obs[i, "site.of.metastasis"]
    same.site <- filter(diss.obs[-i,], site.of.metastasis == site) ## remove the sample in question
    random.sample <- sample(rownames(same.site))[1]
    abundances <- rbind(clr_met[sample,1:6], clr_met[random.sample, 1:6])
    
    # Add to large permutation matrix
    diss.perm[sample,zz] <- vegdist(abundances, method = 'euclidean', binary = FALSE,na.rm = TRUE)
  }
} 


# Plot distribution of observed and distribution of permuted dissimilarity scores
diss.perm.long <- reshape2::melt(diss.perm)
diss.perm.long$type <- "permuted"
diss.obs.long <- as.data.frame(diss.obs$diss_to_PT)
colnames(diss.obs.long) <- "value"
diss.obs.long$type <- "observed"
diss.merged <- rbind(diss.perm.long[,2:3], diss.obs.long)

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_histogram(aes(y=..density..), position="identity", alpha=0.3)+scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

ks.test(diss.obs.long$value, diss.perm.long$value, alternative = "greater")

# Plot distribution of permuted means
obs.median <- median(diss.obs.long$value)
perm.medians <- apply(diss.perm, MARGIN = 2, FUN = median)

# Test for normal distribution of permuted medians
shapiro.test(perm.medians)

# Test probability of the observed median to be drawn from the distribution of permuted medians
sig <- pnorm(obs.median, mean = mean(perm.medians), sd = sd(perm.medians))

ggplot(as.data.frame(perm.medians),aes(x=perm.medians,..scaled..)) + 
  geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=median(diss.obs.long$value), linetype="dashed")+theme(legend.position="top") +
  annotate(geom="text", x=3.8, y=1, label=paste0("p = " , round(sig, digits = 4)), color="red")


```

# T cell subtype abundances

### Calculate Aitchison distance between matched samples

For PTs with multiple matched metastatic sites, these sites are pooled before calculating the Aitchison distance between sample pairs.


```{r Tcell-aitchison}
sce_T <- filterSCE(sce_immune_intra, Tcell_subtype != "none")
sce_met <- filterSCE(sce_T,Tissue.Type!="PT")
sce_PT <- filterSCE(sce_T,Tissue.Type=="PT")

ncs_met <- table(sce_met$patient.MetPT, sce_met$Tcell_subtype)
ncs_PT <- table(sce_PT$patient.MetPT, sce_PT$Tcell_subtype)

# Include only samples with > 6 T cells
ncs_met <- ncs_met[rowSums(ncs_met) > 6,]
ncs_PT <- ncs_PT[rowSums(ncs_PT) > 6,]

# Calculate centered log-transform of abundances
fqs_met  <- log(prop.table(ncs_met + 1, 1), 2)
clr_met  <- as.data.frame.matrix(fqs_met-rowMeans(fqs_met))
fqs_PT <- log(prop.table(ncs_PT+ 1, 1), 2)
clr_PT <- as.data.frame.matrix(fqs_PT-rowMeans(fqs_PT))

clr_met$patient.MetPT <- rownames(clr_met)
clr_met <- separate(clr_met, patient.MetPT, into = c("patient", "MetPT"), sep = "_")

clr_PT$patient.MetPT <- rownames(clr_PT)
clr_PT <- separate(clr_PT, patient.MetPT, into = c("patient", "MetPT"), sep = "_")


# Create empty dissimilarity data frame
diss.obs <- as.data.frame(matrix(ncol = 2, nrow = length(rownames(clr_met)), dimnames = list(rownames(clr_met), c("patient.MetPT", "diss_to_PT"))))

# Loop over metastatic tissues and compute dissimilarity to matched PT
for(i in rownames(clr_met)){
 patientID <- clr_met[i,"patient"]
 abundances <- rbind(clr_met[i,], filter(clr_PT, patient == patientID))
 if(nrow(abundances) == 2) {
   diss <- vegdist(abundances[,1:6],method = 'euclidean', binary = FALSE,na.rm = TRUE)
    diss.obs[i,] <- c(i, diss)
 }
}

# Remove NAs (samples for which no PT data is available)
diss.obs <- na.exclude(diss.obs)

diss.obs <- separate(diss.obs, patient.MetPT, into = c("patient", "MetPT"), sep = "_")
diss.obs$diss_to_PT <- as.numeric(diss.obs$diss_to_PT)
diss.obs$site.of.metastasis <- sample_unique$Site.of.metastasis[match(diss.obs$patient, sample_unique$Patient.ID)]

```

### Random matching of Mets to other Mets of the same site

```{r Tcell-permutations}

# Create empty data frame for 1000 permutations
diss.perm <- as.data.frame(matrix(ncol = 1000, nrow = length(rownames(diss.obs)), dimnames = list(rownames(diss.obs), c(1:1000))))


# Loop for permutations
for (zz in 1:1000) {
  
  # Loop over metastatic tissues and compute dissimilarity to random other metastatis from the same site
  for(i in 1:length(rownames(diss.obs))){
    sample <- rownames(diss.obs)[i]
    site <- diss.obs[i, "site.of.metastasis"]
    same.site <- filter(diss.obs[-i,], site.of.metastasis == site) ## remove the sample in question
    random.sample <- sample(rownames(same.site))[1]
    abundances <- rbind(clr_met[sample,1:11], clr_met[random.sample, 1:11])
    
    # Add to large permutation matrix
    diss.perm[sample,zz] <- vegdist(abundances, method = 'euclidean', binary = FALSE,na.rm = TRUE)
  }
} 


# Plot distribution of observed and distribution of permuted dissimilarity scores
diss.perm.long <- reshape2::melt(diss.perm)
diss.perm.long$type <- "permuted"
diss.obs.long <- as.data.frame(diss.obs$diss_to_PT)
colnames(diss.obs.long) <- "value"
diss.obs.long$type <- "observed"
diss.merged <- rbind(diss.perm.long[,2:3], diss.obs.long)

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

ks.test(diss.obs.long$value, diss.perm.long$value, alternative = "greater")

# Plot distribution of permuted means
obs.median <- median(diss.obs.long$value)
perm.medians <- apply(diss.perm, MARGIN = 2, FUN = median)

# Test for normal distribution of permuted medians
shapiro.test(perm.medians)

# Test probability of the observed median to be drawn from the distribution of permuted medians
sig <- pnorm(obs.median, mean = mean(perm.medians), sd = sd(perm.medians))

ggplot(as.data.frame(perm.medians),aes(x=perm.medians,..scaled..)) + 
  geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=median(diss.obs.long$value), linetype="dashed")+theme(legend.position="top") +
  annotate(geom="text", x=4.0, y=1, label=paste0("p = " , round(sig, digits = 6)), color="red")

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_histogram(aes(y=..density..), position="identity", alpha=0.3)+scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

```

# Myeloid cell subtype abundances

### Calculate Aitchison distance between matched samples

For PTs with multiple matched metastatic sites, these sites are pooled before calculating the Aitchison distance between sample pairs.


```{r myeloid-aitchison}
sce_M <- filterSCE(sce_M_intra, M_cluster != "none")
sce_met <- filterSCE(sce_M,Tissue.Type!="PT")
sce_PT <- filterSCE(sce_M,Tissue.Type=="PT")

ncs_met <- table(sce_met$patient.MetPT, sce_met$M_cluster)
ncs_PT <- table(sce_PT$patient.MetPT, sce_PT$M_cluster)

# Include only samples with > 10 myeloid cells
ncs_met <- ncs_met[rowSums(ncs_met) > 10,]
ncs_PT <- ncs_PT[rowSums(ncs_PT) > 10,]

# Calculate centered log-transform of abundances
fqs_met  <- log(prop.table(ncs_met + 1, 1), 2)
clr_met  <- as.data.frame.matrix(fqs_met-rowMeans(fqs_met))
fqs_PT <- log(prop.table(ncs_PT+ 1, 1), 2)
clr_PT <- as.data.frame.matrix(fqs_PT-rowMeans(fqs_PT))

clr_met$patient.MetPT <- rownames(clr_met)
clr_met <- separate(clr_met, patient.MetPT, into = c("patient", "MetPT"), sep = "_")

clr_PT$patient.MetPT <- rownames(clr_PT)
clr_PT <- separate(clr_PT, patient.MetPT, into = c("patient", "MetPT"), sep = "_")


# Create empty dissimilarity data frame
diss.obs <- as.data.frame(matrix(ncol = 2, nrow = length(rownames(clr_met)), dimnames = list(rownames(clr_met), c("patient.MetPT", "diss_to_PT"))))

# Loop over metastatic tissues and compute dissimilarity to matched PT
for(i in rownames(clr_met)){
 patientID <- clr_met[i,"patient"]
 abundances <- rbind(clr_met[i,], filter(clr_PT, patient == patientID))
 if(nrow(abundances) == 2) {
   diss <- vegdist(abundances[,1:6],method = 'euclidean', binary = FALSE,na.rm = TRUE)
    diss.obs[i,] <- c(i, diss)
 }
}

# Remove NAs (samples for which no PT data is available)
diss.obs <- na.exclude(diss.obs)

diss.obs <- separate(diss.obs, patient.MetPT, into = c("patient", "MetPT"), sep = "_")
diss.obs$diss_to_PT <- as.numeric(diss.obs$diss_to_PT)
diss.obs$site.of.metastasis <- sample_unique$Site.of.metastasis[match(diss.obs$patient, sample_unique$Patient.ID)]
```

```{r myeloid-permutations}

# Create empty data frame for 1000 permutations
diss.perm <- as.data.frame(matrix(ncol = 1000, nrow = length(rownames(diss.obs)), dimnames = list(rownames(diss.obs), c(1:1000))))


# Loop for permutations
for (zz in 1:1000) {
  
  # Loop over metastatic tissues and compute dissimilarity to random other metastatis from the same site
  for(i in 1:length(rownames(diss.obs))){
    sample <- rownames(diss.obs)[i]
    site <- diss.obs[i, "site.of.metastasis"]
    same.site <- filter(diss.obs[-i,], site.of.metastasis == site) ## remove the sample in question
    random.sample <- sample(rownames(same.site))[1]
    abundances <- rbind(clr_met[sample,1:20], clr_met[random.sample, 1:20])
    
    # Add to large permutation matrix
    diss.perm[sample,zz] <- vegdist(abundances, method = 'euclidean', binary = FALSE,na.rm = TRUE)
  }
} 


# Plot distribution of observed and distribution of permuted dissimilarity scores
diss.perm.long <- reshape2::melt(diss.perm)
diss.perm.long$type <- "permuted"
diss.obs.long <- as.data.frame(diss.obs$diss_to_PT)
colnames(diss.obs.long) <- "value"
diss.obs.long$type <- "observed"
diss.merged <- rbind(diss.perm.long[,2:3], diss.obs.long)

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

ks.test(diss.obs.long$value, diss.perm.long$value, alternative = "greater")

# Plot distribution of permuted means
obs.median <- median(diss.obs.long$value)
perm.medians <- apply(diss.perm, MARGIN = 2, FUN = median)

# Test for normal distribution of permuted medians
shapiro.test(perm.medians)

# Test probability of the observed median to be drawn from the distribution of permuted medians
sig <- pnorm(obs.median, mean = mean(perm.medians), sd = sd(perm.medians))

ggplot(as.data.frame(perm.medians),aes(x=perm.medians,..scaled..)) + 
  geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=median(diss.obs.long$value), linetype="dashed")+theme(legend.position="top") +
  annotate(geom="text", x=4.0, y=1, label=paste0("p = " , round(sig, digits = 6)), color="red")

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_histogram(aes(y=..density..), position="identity", alpha=0.3)+scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")
```

