---
title: "ImmuneCombined_02_DA"
author: "SandraTietscher"
date: "2021-12-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(cowplot)
library(tidyr)
library(dplyr)
library(ggpubr)
library(CATALYST)
library(viridis)
library(ggpubr)
library(edgeR)

# Read immune sce
sce_immune <- readRDS("output/SCEs/Tcell/TcellSCE_subclustered.rds")
sce_immune$patient.tissue <- paste(sce_immune$Patient.ID, sce_immune$Tissue.Type, sep = "_")
sce_immune$MetPT <- ifelse(sce_immune$Tissue.Type == "PT", "PT", "Met")
sce_immune$patient.tissue.loc <- paste(sce_immune$patient.tissue, sce_immune$Location, sep = "_")

# Read macrophage sce
sce_M <- readRDS("output/SCEs/myeloid/myeloidSCE_M.rds")
sce_M$patient.tissue <- paste(sce_M$Patient.ID, sce_M$Tissue.Type, sep = "_")
sce_M$MetPT <- ifelse(sce_M$Tissue.Type == "PT", "PT", "Met")
sce_M$patient.tissue.loc <- paste(sce_M$patient.tissue, sce_M$Location, sep = "_")

```

# Plot primary tumor and metastais PCA plots for general immune cells, T cells and myeloid cells
```{r pca-immune-phenotypes}
#Myeloid cells
sce_M$sample_type[sce_M$Tissue.Type == 'PT'] <- "PT"
sce_M$sample_type[sce_M$Tissue.Type == 'Liver'] <- "M"
sce_M$sample_type[sce_M$Tissue.Type == 'Bone'] <- "M"
sce_M$sample_type[sce_M$Tissue.Type == 'Brain'] <- "M"
sce_M$sample_type[sce_M$Tissue.Type == 'SoftTissue'] <- "M"
sce_M$sample_type <- factor(sce_M$sample_type)
sce_M$sample_id <- factor(paste(sce_M$Patient.ID, sce_M$sample_type)) 
sce_M$cluster_id <- sce_M$M_cluster

metadata(sce_M)$cluster_codes <- data.frame(custom = factor(levels(sce_M$cluster_id)), levels = levels(sce_M$cluster_id))
sce_M$condition <- factor(sce_M$Tissue.Type, levels = c( "PT", "Bone", "Brain","Liver","SoftTissue"))
clrDR(sce_M, by = "sample_id", k = "custom",size_by = FALSE,point_pal= c("#4C9900","#FF8000","#AC16AC","#FF3333","#CCCC00"), arrows= TRUE)


#Immune cells
sce_immune$sample_type[sce_immune$Tissue.Type == 'PT'] <- "PT"
sce_immune$sample_type[sce_immune$Tissue.Type == 'Liver'] <- "M"
sce_immune$sample_type[sce_immune$Tissue.Type == 'Bone'] <- "M"
sce_immune$sample_type[sce_immune$Tissue.Type == 'Brain'] <- "M"
sce_immune$sample_type[sce_immune$Tissue.Type == 'SoftTissue'] <- "M"
sce_immune$sample_type <- factor(sce_immune$sample_type)
sce_immune$sample_id <- factor(paste(sce_immune$Patient.ID, sce_immune$sample_type))
sce_immune$condition <- factor(sce_immune$Tissue.Type, levels = c( "PT", "Bone", "Brain","Liver","SoftTissue"))

# Subset object
sce_immune <- sce_immune[,which(sce_immune$cell_class == "immune")]
# Exclude HLA-DR_high cells (undefined cell class)
sce_immune <- sce_immune[,which(sce_immune$celltype != "HLA-DR_high")]
# Classify neutrophils as myeloid cells
sce_immune$celltype <- ifelse(sce_immune$celltype == "neutrophil", "myeloid", sce_immune$celltype)

sce_immune$cluster_id <- as.factor(sce_immune$celltype)
metadata(sce_immune)$cluster_codes <- data.frame(custom = factor(levels(sce_immune$cluster_id)), levels = levels(sce_immune$cluster_id))
clrDR(sce_immune, by = "sample_id", k = "custom",size_by = FALSE,point_pal= c("#4C9900","#FF8000","#AC16AC","#FF3333","#CCCC00"), arrows= TRUE)


# Subset object
sce_T <- sce_immune[,which(sce_immune$Tcell_subtype != "none")]
sce_T$cluster_id <- as.factor(sce_T$Tcell_subtype)
metadata(sce_T)$cluster_codes <- data.frame(custom = factor(levels(sce_T$cluster_id)), levels = levels(sce_T$cluster_id))

sce_T$sample_type[sce_T$Tissue.Type == 'PT'] <- "PT"
sce_T$sample_type[sce_T$Tissue.Type == 'Liver'] <- "M"
sce_T$sample_type[sce_T$Tissue.Type == 'Bone'] <- "M"
sce_T$sample_type[sce_T$Tissue.Type == 'Brain'] <- "M"
sce_T$sample_type[sce_T$Tissue.Type == 'SoftTissue'] <- "M"
sce_T$sample_type <- factor(sce_T$sample_type)
sce_T$sample_id <- factor(paste(sce_T$Patient.ID, sce_T$sample_type))

clrDR(sce_T, by = "sample_id", k = "custom",size_by = FALSE,point_pal= c("#4C9900","#FF8000","#AC16AC","#FF3333","#CCCC00"), arrows= TRUE)
```







# All locations: Differential abundance of cell types in PT vs met metastasis 

Patient matching is accounted for by adding Patient ID to the design matrix for the model.
Location is accounted for by adding it as an additional blocking factor to the design matrix.

### General immune cells

Over all metastatic tissues combined

```{r immuneCells-allSites, fig.width=4.5, fig.height=3}
# Subset object
sce_immune <- sce_immune[,which(sce_immune$cell_class == "immune")]

# Exclude HLA-DR_high cells (undefined cell class)
sce_immune <- sce_immune[,which(sce_immune$celltype != "HLA-DR_high")]

# Classify neutrophils as myeloid cells
sce_immune$celltype <- ifelse(sce_immune$celltype == "neutrophil", "myeloid", sce_immune$celltype)

# Prepare abundance object
abundances <- table(sce_immune$celltype, sce_immune$patient.tissue.loc) 
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_immune)[match(colnames(abundances), sce_immune$patient.tissue.loc),]
MetPT <- factor(meta$MetPT)
meta$MetPT <- relevel(x=MetPT, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$MetPT)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Location + MetPT, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=colnames(fit.ab)[ncol(fit.ab)])
summary(decideTests(res))
DA <- topTags(res)$table

DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

# Sort by logFC
order <- rownames(DA[order(DA$logFC),])
DA <- DA[order,]
DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("All met vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For each metastatic tissue separately.

```{r immuneCells-sepSites, fig.width=9, fig.height=5}

# Relevel so that PT is the baseline group
Tissue.Type <- factor(meta$Tissue.Type)
meta$Tissue.Type <- relevel(x=Tissue.Type, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$Tissue.Type)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Location + Tissue.Type, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
p.list <- list()

for (i in colnames(fit.ab)[(ncol(fit.ab)-3):ncol(fit.ab)]) {

  res <- glmQLFTest(fit.ab, coef=i)
  summary(decideTests(res))
  DA <- topTags(res)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

p.list[[i]] <- ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
}

plot_grid(plotlist = p.list, ncol = 2)

```

### T cell subtypes

Over all metastatic tissues combined

```{r Tcells-allSites, fig.width=4.5, fig.height=3}
# Subset object
sce_T <- sce_immune[,which(sce_immune$Tcell_subtype != "none")]

# Prepare abundance object
abundances <- table(sce_T$Tcell_subtype, sce_T$patient.tissue.loc) 
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_T)[match(colnames(abundances), sce_T$patient.tissue.loc),]
MetPT <- factor(meta$MetPT)
meta$MetPT <- relevel(x=MetPT, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$MetPT)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Location + MetPT, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=colnames(fit.ab)[ncol(fit.ab)])
summary(decideTests(res))
DA <- topTags(res, n = 15)$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

# Sort by logFC
order <- rownames(DA[order(DA$logFC),])
DA <- DA[order,]
DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("All met vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For each metastatic tissue separately.

```{r Tcells-sepSites, fig.width=9, fig.height=5}
Tissue.Type <- factor(meta$Tissue.Type)
meta$Tissue.Type <- relevel(x=Tissue.Type, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$Tissue.Type)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Location + Tissue.Type, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
p.list <- list()

for (i in colnames(fit.ab)[(ncol(fit.ab)-3):ncol(fit.ab)]) {

  res <- glmQLFTest(fit.ab, coef=i)
  summary(decideTests(res))
  DA <- topTags(res, n = 15)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

p.list[[i]] <- ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
}

plot_grid(plotlist = p.list, ncol = 2)

```


### Myeloid cell subtypes

Over all metastatic tissues combined

```{r myeloid-allSites, fig.width=4.5, fig.height=3}
# Subset object
sce_M <- sce_M[,which(sce_M$M_cluster != "none")]

# Prepare abundance object
abundances <- table(sce_M$M_cluster, sce_M$patient.tissue.loc) 
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_M)[match(colnames(abundances), sce_M$patient.tissue.loc),]
MetPT <- factor(meta$MetPT)
meta$MetPT <- relevel(x=MetPT, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$MetPT)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Location + MetPT, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=colnames(fit.ab)[ncol(fit.ab)])
summary(decideTests(res))
DA <- topTags(res, n = 25)$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

# Sort by logFC
order <- rownames(DA[order(DA$logFC),])
DA <- DA[order,]
DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("All met vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For each metastatic tissue separately.

```{r myeloid-sepSites, fig.width=9, fig.height=5}

Tissue.Type <- factor(meta$Tissue.Type)
meta$Tissue.Type <- relevel(x=Tissue.Type, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$Tissue.Type)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Location + Tissue.Type, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
p.list <- list()

for (i in colnames(fit.ab)[(ncol(fit.ab)-3):ncol(fit.ab)]) {

  res <- glmQLFTest(fit.ab, coef=i)
  summary(decideTests(res))
  DA <- topTags(res, n = 25)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

p.list[[i]] <- ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
}

plot_grid(plotlist = p.list, ncol = 2)

```


# Intratumoral only locations: Differential abundance of cell types in PT vs met metastasis

Only intratumoral images are used to prevent location bias.

### General immune cells

```{r immune-intratumoral-allSites, fig.width=4.5, fig.height=3}
# Subset object
sce_immune_intra <- sce_immune[,which(sce_immune$Location == "intratumoral")]

# Prepare abundance object
abundances <- table(sce_immune_intra$celltype, sce_immune_intra$patient.tissue.loc) 
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_immune_intra)[match(colnames(abundances), sce_immune_intra$patient.tissue.loc),]
MetPT <- factor(meta$MetPT)
meta$MetPT <- relevel(x=MetPT, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$MetPT)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + MetPT, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=colnames(fit.ab)[ncol(fit.ab)])
summary(decideTests(res))
DA <- topTags(res)$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

# Sort by logFC
order <- rownames(DA[order(DA$logFC),])
DA <- DA[order,]
DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("All met vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For each metastatic tissue separately.

```{r immune-intratumoral-sepSites}
Tissue.Type <- factor(meta$Tissue.Type)
meta$Tissue.Type <- relevel(x=Tissue.Type, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$Tissue.Type)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Tissue.Type, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
p.list <- list()

for (i in colnames(fit.ab)[(ncol(fit.ab)-3):ncol(fit.ab)]) {

  res <- glmQLFTest(fit.ab, coef=i)
  summary(decideTests(res))
  DA <- topTags(res)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

p.list[[i]] <- ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
}

plot_grid(plotlist = p.list, ncol = 2)
```


### T cell subtypes

Over all metastatic tissues combined

```{r Tcells-intratumoral-allSites, fig.width=4.5, fig.height=3}
# Subset object
sce_T_intra <- sce_immune_intra[,which(sce_immune_intra$Tcell_subtype != "none")]

# Prepare abundance object
abundances <- table(sce_T_intra$Tcell_subtype, sce_T_intra$patient.tissue.loc) 
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_T_intra)[match(colnames(abundances), sce_T_intra$patient.tissue.loc),]
MetPT <- factor(meta$MetPT)
meta$MetPT <- relevel(x=MetPT, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$MetPT)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + MetPT, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=colnames(fit.ab)[ncol(fit.ab)])
summary(decideTests(res))
DA <- topTags(res, n=15)$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

# Sort by logFC
order <- rownames(DA[order(DA$logFC),])
DA <- DA[order,]
DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("All met vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For each metastatic tissue separately.

```{r Tcell-intratumoral-sepSites, fig.width=9, fig.height=5}
Tissue.Type <- factor(meta$Tissue.Type)
meta$Tissue.Type <- relevel(x=Tissue.Type, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$Tissue.Type)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Tissue.Type, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
p.list <- list()

for (i in colnames(fit.ab)[(ncol(fit.ab)-3):ncol(fit.ab)]) {

  res <- glmQLFTest(fit.ab, coef=i)
  summary(decideTests(res))
  DA <- topTags(res, n=15)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

p.list[[i]] <- ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
}

plot_grid(plotlist = p.list, ncol = 2)

```


### Myeloid cell subtypes

Over all metastatic tissues combined

```{r myeloid-intratumoral-allSites, fig.width=4.5, fig.height=3}
# Subset object
sce_M_intra <- sce_M[,which(sce_M$Location == "intratumoral")]

# Prepare abundance object
abundances <- table(sce_M_intra$M_cluster, sce_M_intra$patient.tissue.loc) 
abundances <- unclass(abundances) 

# Attach column metadata (Relevel so that PT is the baseline group)
meta <- colData(sce_M_intra)[match(colnames(abundances), sce_M_intra$patient.tissue.loc),]
MetPT <- factor(meta$MetPT)
meta$MetPT <- relevel(x=MetPT, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$MetPT)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + MetPT, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
res <- glmQLFTest(fit.ab, coef=colnames(fit.ab)[ncol(fit.ab)])
summary(decideTests(res))
DA <- topTags(res, n = 25)$table
DA$celltype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")

# Sort by logFC
order <- rownames(DA[order(DA$logFC),])
DA <- DA[order,]
DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("All met vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For each metastatic tissue separately.

```{r myeloid-intratumoral-sepSites, fig.width=9, fig.height=5}
Tissue.Type <- factor(meta$Tissue.Type)
meta$Tissue.Type <- relevel(x=Tissue.Type, ref = "PT")
y.ab <- DGEList(abundances, samples=meta, group = meta$Tissue.Type)

# Define the design matrix for the model: Add Patient ID to the model to account for matched design, add Location to account for different tumor regions, add tissue type as an additive factor
design <- model.matrix(~Patient.ID + Tissue.Type, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance of cell types in PT vs met tissue and plot
p.list <- list()

for (i in colnames(fit.ab)[(ncol(fit.ab)-3):ncol(fit.ab)]) {

  res <- glmQLFTest(fit.ab, coef=i)
  summary(decideTests(res))
  DA <- topTags(res, n = 25)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

p.list[[i]] <- ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0(i, "vs PT"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
}

plot_grid(plotlist = p.list, ncol = 2)

```


