---
title: "ImmuneCombined_03_dissimilarity"
author: "SandraTietscher"
date: "2021-12-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


# Load libraries and data
```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(dittoSeq)
library(scales)
library(tidyverse)
library(ggpubr)
library(cowplot)
library(scran)
library(CATALYST)
library(edgeR)
library(vegan)
library(ComplexHeatmap)


# Read immune sce
sce_immune <- readRDS("output/SCEs/Tcell/TcellSCE_subclustered.rds")
sce_immune$patient.tissue <- paste(sce_immune$Patient.ID, sce_immune$Tissue.Type, sep = "_")
sce_immune$MetPT <- ifelse(sce_immune$Tissue.Type == "PT", "PT", "Met")
sce_immune$patient.MetPT <- paste(sce_immune$Patient.ID, sce_immune$MetPT, sep = "_")

# Read macrophage sce
sce_M <- readRDS("output/SCEs/myeloid/myeloidSCE_M.rds")
sce_M$patient.tissue <- paste(sce_M$Patient.ID, sce_M$Tissue.Type, sep = "_")
sce_M$MetPT <- ifelse(sce_M$Tissue.Type == "PT", "PT", "Met")
sce_M$patient.MetPT <- paste(sce_M$Patient.ID, sce_M$MetPT, sep = "_")

# Read in ROI metadata for immune panel
ROI_meta <- read.csv("data/immune_images_patient_metadata_encoded.csv")
ROI_meta$patient.tissue <- paste(ROI_meta$Patient.ID, ROI_meta$Tissue.Type, sep = "_")
sample_unique <- ROI_meta[!duplicated(ROI_meta$patient.tissue),]
rownames(sample_unique) <- sample_unique$patient.tissue


# Read in ROI metadata for myeloid panel
ROI_meta_M <- read.csv("data/macrophage_images_patient_metadata_encoded.csv")

ROI_meta_M$patient.tissue <- paste(ROI_meta_M$Patient.ID, ROI_meta_M$Tissue.Type, sep = "_")
sample_unique_M <- ROI_meta_M[!duplicated(ROI_meta_M$patient.tissue),]
rownames(sample_unique_M) <- sample_unique_M$patient.tissue

```

# Subset: Use only intratumoral samples

```{r subset-intratumoral}
sce_immune_intra <- filterSCE(sce_immune, Location == "intratumoral")
sce_M_intra <- filterSCE(sce_M, Location == "intratumoral")
```


# General immune cell abundances

### Calculate Aitchison distance between matched samples

For PTs with multiple matched metastatic sites, these sites are pooled before calculating the Aitchison distance between sample pairs.

```{r immune-aitchison}
sce_im <- filterSCE(sce_immune_intra, celltype %in% c("NK_cell", "T_cell", "myeloid", "plasma_cell", "B_cell", "neutrophil"))
sce_met <- filterSCE(sce_im,Tissue.Type!="PT")
sce_PT <- filterSCE(sce_im,Tissue.Type=="PT")

# Calculate centered log-transform of abundances
ncs_met <- table(sce_met$patient.MetPT, sce_met$celltype)
fqs_met  <- log(prop.table(ncs_met + 1, 1), 2)
clr_met  <- as.data.frame.matrix(fqs_met-rowMeans(fqs_met))

ncs_PT <- table(sce_PT$patient.MetPT, sce_PT$celltype)
fqs_PT <- log(prop.table(ncs_PT+ 1, 1), 2)
clr_PT <- as.data.frame.matrix(fqs_PT-rowMeans(fqs_PT))

clr_met$patient.MetPT <- rownames(clr_met)
clr_met <- separate(clr_met, patient.MetPT, into = c("patient", "MetPT"), sep = "_")

clr_PT$patient.MetPT <- rownames(clr_PT)
clr_PT <- separate(clr_PT, patient.MetPT, into = c("patient", "MetPT"), sep = "_")


# Create empty dissimilarity data frame
diss.obs <- as.data.frame(matrix(ncol = 2, nrow = length(rownames(clr_met)), dimnames = list(rownames(clr_met), c("patient.MetPT", "diss_to_PT"))))

# Loop over metastatic tissues and compute dissimilarity to matched PT
for(i in rownames(clr_met)){
 patientID <- clr_met[i,"patient"]
 abundances <- rbind(clr_met[i,], filter(clr_PT, patient == patientID))
 if(nrow(abundances) == 2) {
   diss <- vegdist(abundances[,1:6],method = 'euclidean', binary = FALSE,na.rm = TRUE)
    diss.obs[i,] <- c(i, diss)
 }
}

# Remove NAs (samples for which no PT data is available)
diss.obs <- na.exclude(diss.obs)

# Check correlation of dissimilarity with number of cells
diss.obs <- separate(diss.obs, patient.MetPT, into = c("patient", "MetPT"), sep = "_")
diss.obs$diss_to_PT <- as.numeric(diss.obs$diss_to_PT)
diss.obs$cellnr_met <- table(sce_met$patient.MetPT)[rownames(diss.obs)]
diss.obs$cellnr_PT <- table(sce_PT$patient.MetPT)[paste0(diss.obs$patient, "_PT")]

ggplot(diss.obs, aes(diss_to_PT, cellnr_met))+
  geom_point()+
  stat_cor(method="spearman", label.x.npc = "middle", color = "black") + 
  geom_smooth(method = lm, color = "black", se=FALSE)+ 
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())

ggplot(diss.obs, aes(diss_to_PT, cellnr_PT))+
  geom_point()+
  stat_cor(method="spearman", label.x.npc = "middle", color = "black") + 
  geom_smooth(method = lm, color = "black", se=FALSE)+ 
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())
```

### Compare dissimilarity for different tissue types

```{r immune-diss-byTissue}
diss.obs$site.of.metastasis <- sample_unique$Site.of.metastasis[match(diss.obs$patient, sample_unique$Patient.ID)]

# Plot as Boxplots
ggplot(diss.obs, aes(site.of.metastasis, diss_to_PT)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.1)+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))+
  stat_compare_means(method = "kruskal.test")

```

### Random permutation of PT-Met matching

```{r immune-permutations}

# Create empty data frame for 10000 permutations
diss.perm <- as.data.frame(matrix(ncol = 10000, nrow = length(rownames(diss.obs)), dimnames = list(rownames(diss.obs), c(1:10000))))

# Subset PTs to include only those with matched metastasis
clr_PT <- filter(clr_PT, patient %in% diss.obs$patient)

# Loop for permutations
for (zz in 1:10000) {
 
  PTs_random <- sample(rownames(clr_PT), size = length(rownames(clr_PT)))
  
  # Loop over metastatic tissues and compute dissimilarity to RANDOM PT
  for(i in 1:length(rownames(diss.obs))){
    Met <- rownames(diss.obs)[i]
    PT <- PTs_random[i]
    abundances <- rbind(clr_met[Met,1:6], clr_PT[PT, 1:6])
    
    # Add to large permutation matrix
    diss.perm[Met,zz] <- vegdist(abundances, method = 'euclidean', binary = FALSE,na.rm = TRUE)
  }
} 


# Plot distribution of observed and distribution of permuted dissimilarity scores
diss.perm.long <- reshape2::melt(diss.perm)
diss.perm.long$type <- "permuted"
diss.obs.long <- as.data.frame(diss.obs$diss_to_PT)
colnames(diss.obs.long) <- "value"
diss.obs.long$type <- "observed"
diss.merged <- rbind(diss.perm.long[,2:3], diss.obs.long)

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

ks.test(diss.obs.long$value, diss.perm.long$value, alternative = "greater")

# Plot distribution of permuted means
obs.median <- median(diss.obs.long$value)
perm.medians <- apply(diss.perm, MARGIN = 2, FUN = median)

# Test for normal distribution of permuted medians
shapiro.test(perm.medians)

# Test probability of the observed median to be drawn from the distribution of permuted medians
sig <- pnorm(obs.median, mean = mean(perm.medians), sd = sd(perm.medians))

ggplot(as.data.frame(perm.medians),aes(x=perm.medians,..scaled..)) + 
  geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=median(diss.obs.long$value), linetype="dashed")+theme(legend.position="top") +
  annotate(geom="text", x=3.8, y=1, label=paste0("p = " , round(sig, digits = 4)), color="red")
```

# T cell subtype abundances

### Calculate Aitchison distance between matched samples

For PTs with multiple matched metastatic sites, these sites are pooled before calculating the Aitchison distance between sample pairs.


```{r Tcell-aitchison}
sce_T <- filterSCE(sce_immune_intra, Tcell_subtype != "none")
sce_met <- filterSCE(sce_T,Tissue.Type!="PT")
sce_PT <- filterSCE(sce_T,Tissue.Type=="PT")

ncs_met <- table(sce_met$patient.MetPT, sce_met$Tcell_subtype)
ncs_PT <- table(sce_PT$patient.MetPT, sce_PT$Tcell_subtype)

# Include only samples with > 6 T cells
ncs_met <- ncs_met[rowSums(ncs_met) > 6,]
ncs_PT <- ncs_PT[rowSums(ncs_PT) > 6,]

# Calculate centered log-transform of abundances
fqs_met  <- log(prop.table(ncs_met + 1, 1), 2)
clr_met  <- as.data.frame.matrix(fqs_met-rowMeans(fqs_met))
fqs_PT <- log(prop.table(ncs_PT+ 1, 1), 2)
clr_PT <- as.data.frame.matrix(fqs_PT-rowMeans(fqs_PT))

clr_met$patient.MetPT <- rownames(clr_met)
clr_met <- separate(clr_met, patient.MetPT, into = c("patient", "MetPT"), sep = "_")

clr_PT$patient.MetPT <- rownames(clr_PT)
clr_PT <- separate(clr_PT, patient.MetPT, into = c("patient", "MetPT"), sep = "_")


# Create empty dissimilarity data frame
diss.obs <- as.data.frame(matrix(ncol = 2, nrow = length(rownames(clr_met)), dimnames = list(rownames(clr_met), c("patient.MetPT", "diss_to_PT"))))

# Loop over metastatic tissues and compute dissimilarity to matched PT
for(i in rownames(clr_met)){
 patientID <- clr_met[i,"patient"]
 abundances <- rbind(clr_met[i,], filter(clr_PT, patient == patientID))
 if(nrow(abundances) == 2) {
   diss <- vegdist(abundances[,1:6],method = 'euclidean', binary = FALSE,na.rm = TRUE)
    diss.obs[i,] <- c(i, diss)
 }
}

# Remove NAs (samples for which no PT data is available)
diss.obs <- na.exclude(diss.obs)

# Check correlation of dissimilarity with number of cells
diss.obs <- separate(diss.obs, patient.MetPT, into = c("patient", "MetPT"), sep = "_")
diss.obs$diss_to_PT <- as.numeric(diss.obs$diss_to_PT)
diss.obs$cellnr_met <- table(sce_met$patient.MetPT)[rownames(diss.obs)]
diss.obs$cellnr_PT <- table(sce_PT$patient.MetPT)[paste0(diss.obs$patient, "_PT")]

ggplot(diss.obs, aes(diss_to_PT, cellnr_met))+
  geom_point()+
  stat_cor(method="spearman", label.x.npc = "middle", color = "black") + 
  geom_smooth(method = lm, color = "black", se=FALSE)+ 
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())

ggplot(diss.obs, aes(diss_to_PT, cellnr_PT))+
  geom_point()+
  stat_cor(method="spearman", label.x.npc = "middle", color = "black") + 
  geom_smooth(method = lm, color = "black", se=FALSE)+ 
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())
```

### Compare dissimilarity for different tissue types

```{r Tcell-diss-byTissue}

diss.obs$site.of.metastasis <- sample_unique$Site.of.metastasis[match(diss.obs$patient, sample_unique$Patient.ID)]

# Plot as Boxplots
ggplot(diss.obs, aes(site.of.metastasis, diss_to_PT)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.1)+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))+
  stat_compare_means(method = "kruskal.test")
```

### Example barplots for samples that are very similar and very dissimilar to matched PT

```{r example-barplots}
tab.bc <- prop.table(table(sce_T$patient.tissue, sce_T$Tcell_subtype), margin = 1)

# Similar examples
ID36_brain <- tab.bc[rownames(tab.bc) %in% c("ID36_Brain", "ID36_PT"),]
ID9_bone <- tab.bc[rownames(tab.bc) %in% c("ID9_Bone", "ID9_PT"),]
ID60_softtissue <- tab.bc[rownames(tab.bc) %in% c("ID60_SoftTissue", "ID60_PT"),]
ID23_liver <- tab.bc[rownames(tab.bc) %in% c("ID23_Liver", "ID23_PT"),]

# Dissimilar examples
ID78_brain <- tab.bc[rownames(tab.bc) %in% c("ID78_Brain", "ID78_PT"),]
ID10_bone <- tab.bc[rownames(tab.bc) %in% c("ID10_Bone", "ID10_PT"),]
ID4_softtissue <- tab.bc[rownames(tab.bc) %in% c("ID4_SoftTissue", "ID4_PT"),]
ID68_liver <- tab.bc[rownames(tab.bc) %in% c("ID68_Liver", "ID68_PT"),]


comb <- as.data.frame(rbind(ID36_brain, ID9_bone, ID23_liver, ID10_bone, ID78_brain, ID68_liver, ID60_softtissue, ID4_softtissue))
comb$patient.tissue <- rownames(comb)
comb <- separate(comb, patient.tissue, into = c("patient", "tissue"))
comb.long <- pivot_longer(comb, 1:11, names_to = "Tcell_subtype", values_to = "proportion")
comb.long$tissue <- ifelse(comb.long$tissue == "PT", "PT", "Met")
comb.long$patient <- factor(comb.long$patient, levels = c("ID36", "ID9", "ID60", "ID23", "ID78", "ID10", "ID4", "ID68"))

#Define colors
T_col <- c("#666600", "#003319", "#CCCC00", "#00CC00", "#00FF00", "#B2FF66", "#CCFFCC", "#66FFFF", "#00CCCC", "#3333FF", "#99CCFF")
names(T_col) <- unique(comb.long$Tcell_subtype)

#Plot
ggplot(comb.long, aes(tissue, proportion, fill = Tcell_subtype))+
  geom_col()+
  scale_fill_manual(values = T_col)+
  facet_wrap(~patient, ncol = 4)+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        panel.background = element_blank())
```

### Random permutation of PT-Met matching

```{r Tcell-permutations}

# Create empty data frame for 10000 permutations
diss.perm <- as.data.frame(matrix(ncol = 10000, nrow = length(rownames(diss.obs)), dimnames = list(rownames(diss.obs), c(1:10000))))

# Subset PTs to include only those with matched metastasis
clr_PT <- filter(clr_PT, patient %in% diss.obs$patient)

# Loop for permutations
for (zz in 1:10000) {
 
  PTs_random <- sample(rownames(clr_PT), size = length(rownames(clr_PT)))
  
  # Loop over metastatic tissues and compute dissimilarity to RANDOM PT
  for(i in 1:length(rownames(diss.obs))){
    Met <- rownames(diss.obs)[i]
    PT <- PTs_random[i]
    abundances <- rbind(clr_met[Met,1:6], clr_PT[PT, 1:6])
    
    # Add to large permutation matrix
    diss.perm[Met,zz] <- vegdist(abundances, method = 'euclidean', binary = FALSE,na.rm = TRUE)
  }
} 


# Plot distribution of observed and distribution of permuted dissimilarity scores
diss.perm.long <- reshape2::melt(diss.perm)
diss.perm.long$type <- "permuted"
diss.obs.long <- as.data.frame(diss.obs$diss_to_PT)
colnames(diss.obs.long) <- "value"
diss.obs.long$type <- "observed"
diss.merged <- rbind(diss.perm.long[,2:3], diss.obs.long)

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

ks.test(diss.obs.long$value, diss.perm.long$value, alternative = "greater")

# Plot distribution of permuted means
obs.median <- median(diss.obs.long$value)
perm.medians <- apply(diss.perm, MARGIN = 2, FUN = median)

# Test for normal distribution of permuted medians
shapiro.test(perm.medians)

# Test probability of the observed median to be drawn from the distribution of permuted medians
sig <- pnorm(obs.median, mean = mean(perm.medians), sd = sd(perm.medians))

ggplot(as.data.frame(perm.medians),aes(x=perm.medians,..scaled..)) + 
  geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=median(diss.obs.long$value), linetype="dashed")+theme(legend.position="top") +
  annotate(geom="text", x=3.8, y=1, label=paste0("p = " , round(sig, digits = 4)), color="red")
```


# Myeloid cell subtype abundances

### Calculate Aitchison distance between matched samples

For PTs with multiple matched metastatic sites, these sites are pooled before calculating the Aitchison distance between sample pairs.


```{r myeloid-aitchison}
sce_M <- filterSCE(sce_M_intra, M_cluster != "none")
sce_met <- filterSCE(sce_M,Tissue.Type!="PT")
sce_PT <- filterSCE(sce_M,Tissue.Type=="PT")

ncs_met <- table(sce_met$patient.MetPT, sce_met$M_cluster)
ncs_PT <- table(sce_PT$patient.MetPT, sce_PT$M_cluster)

# Include only samples with > 10 myeloid cells
ncs_met <- ncs_met[rowSums(ncs_met) > 10,]
ncs_PT <- ncs_PT[rowSums(ncs_PT) > 10,]

# Calculate centered log-transform of abundances
fqs_met  <- log(prop.table(ncs_met + 1, 1), 2)
clr_met  <- as.data.frame.matrix(fqs_met-rowMeans(fqs_met))
fqs_PT <- log(prop.table(ncs_PT+ 1, 1), 2)
clr_PT <- as.data.frame.matrix(fqs_PT-rowMeans(fqs_PT))

clr_met$patient.MetPT <- rownames(clr_met)
clr_met <- separate(clr_met, patient.MetPT, into = c("patient", "MetPT"), sep = "_")

clr_PT$patient.MetPT <- rownames(clr_PT)
clr_PT <- separate(clr_PT, patient.MetPT, into = c("patient", "MetPT"), sep = "_")


# Create empty dissimilarity data frame
diss.obs <- as.data.frame(matrix(ncol = 2, nrow = length(rownames(clr_met)), dimnames = list(rownames(clr_met), c("patient.MetPT", "diss_to_PT"))))

# Loop over metastatic tissues and compute dissimilarity to matched PT
for(i in rownames(clr_met)){
 patientID <- clr_met[i,"patient"]
 abundances <- rbind(clr_met[i,], filter(clr_PT, patient == patientID))
 if(nrow(abundances) == 2) {
   diss <- vegdist(abundances[,1:6],method = 'euclidean', binary = FALSE,na.rm = TRUE)
    diss.obs[i,] <- c(i, diss)
 }
}

# Remove NAs (samples for which no PT data is available)
diss.obs <- na.exclude(diss.obs)

# Check correlation of dissimilarity with number of cells
diss.obs <- separate(diss.obs, patient.MetPT, into = c("patient", "MetPT"), sep = "_")
diss.obs$diss_to_PT <- as.numeric(diss.obs$diss_to_PT)
diss.obs$cellnr_met <- table(sce_met$patient.MetPT)[rownames(diss.obs)]
diss.obs$cellnr_PT <- table(sce_PT$patient.MetPT)[paste0(diss.obs$patient, "_PT")]

ggplot(diss.obs, aes(diss_to_PT, cellnr_met))+
  geom_point()+
  stat_cor(method="spearman", label.x.npc = "middle", color = "black") + 
  geom_smooth(method = lm, color = "black", se=FALSE)+ 
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())

ggplot(diss.obs, aes(diss_to_PT, cellnr_PT))+
  geom_point()+
  stat_cor(method="spearman", label.x.npc = "middle", color = "black") + 
  geom_smooth(method = lm, color = "black", se=FALSE)+ 
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.background = element_blank())
```

### Compare dissimilarity for different tissue types

```{r myeloid-diss-byTissue}
diss.obs$site.of.metastasis <- sample_unique$Site.of.metastasis[match(diss.obs$patient, sample_unique$Patient.ID)]

# Plot as Boxplots
ggplot(diss.obs, aes(site.of.metastasis, diss_to_PT)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width = 0.1)+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))+
  stat_compare_means(method = "kruskal.test")
```

### Random permutation of PT-Met matching

```{r myeloid-permutations}

# Create empty data frame for 10000 permutations
diss.perm <- as.data.frame(matrix(ncol = 10000, nrow = length(rownames(diss.obs)), dimnames = list(rownames(diss.obs), c(1:10000))))

# Subset PTs to include only those with matched metastasis
clr_PT <- filter(clr_PT, patient %in% diss.obs$patient)

# Loop for permutations
for (zz in 1:10000) {
 
  PTs_random <- sample(rownames(clr_PT), size = length(rownames(clr_PT)))
  
  # Loop over metastatic tissues and compute dissimilarity to RANDOM PT
  for(i in 1:length(rownames(diss.obs))){
    Met <- rownames(diss.obs)[i]
    PT <- PTs_random[i]
    abundances <- rbind(clr_met[Met,1:6], clr_PT[PT, 1:6])
    
    # Add to large permutation matrix
    diss.perm[Met,zz] <- vegdist(abundances, method = 'euclidean', binary = FALSE,na.rm = TRUE)
  }
} 


# Plot distribution of observed and distribution of permuted dissimilarity scores
diss.perm.long <- reshape2::melt(diss.perm)
diss.perm.long$type <- "permuted"
diss.obs.long <- as.data.frame(diss.obs$diss_to_PT)
colnames(diss.obs.long) <- "value"
diss.obs.long$type <- "observed"
diss.merged <- rbind(diss.perm.long[,2:3], diss.obs.long)

ggplot(diss.merged,aes(x=value, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=mean(diss.obs.long$value, color=type[1]),
             linetype="dashed")+theme(legend.position="top")

ks.test(diss.obs.long$value, diss.perm.long$value, alternative = "greater")

# Plot distribution of permuted means
obs.median <- median(diss.obs.long$value)
perm.medians <- apply(diss.perm, MARGIN = 2, FUN = median)

# Test for normal distribution of permuted medians
shapiro.test(perm.medians)

# Test probability of the observed median to be drawn from the distribution of permuted medians
sig <- pnorm(obs.median, mean = mean(perm.medians), sd = sd(perm.medians))

ggplot(as.data.frame(perm.medians),aes(x=perm.medians,..scaled..)) + 
  geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(xintercept=median(diss.obs.long$value), linetype="dashed")+theme(legend.position="top") +
  annotate(geom="text", x=4.5, y=1, label=paste0("p = " , round(sig, digits = 5)), color="red")
```


# Correlation of mean marker expression between sample pairs

### CD8+ T cells

```{r CD8-marker-correlation}

sce_CD8 <- filterSCE(sce_immune_intra, clustername == "T_CD8")
markers_include_CD8 <- c("CD38", "LAG-3", "CD279", "GITR", "CD7", "Granzyme_B", "TCF7", "CXCL13", "Tim-3", "IRF4")

matched.corr <- as.data.frame(matrix(nrow = 0, ncol = 3))
colnames(matched.corr) <- c("Marker", "Spearman_rho", "Spearman_pvalue")

for (i in markers_include_CD8) {
  i_counts <- as.data.frame(assay(sce_CD8, "counts")[i,])
  colnames(i_counts) <- "i_counts"
  i_counts$patient.MetPT<- sce_CD8$patient.MetPT
  
  i_avg <- i_counts %>% group_by(patient.MetPT) %>% summarize(average = mean(i_counts))
  i_avg <- separate(i_avg, patient.MetPT, into = c("Patient", "Tissue.Class"), sep = "_")
  pat_incl <- table(i_avg$Patient)
  pat_incl <- names(pat_incl[pat_incl > 1])
  i_avg <- filter(i_avg, Patient %in% pat_incl)
  i_avg <- pivot_wider(i_avg, names_from = "Tissue.Class", values_from = "average")
  i_avg$metSite <- ROI_meta$Site.of.metastasis[match(i_avg$Patient, ROI_meta$Patient.ID)]
  i_spearman <- cor.test(i_avg$Met, i_avg$PT, method = "spearman")
  matched.corr[i,] <- c(i, i_spearman$estimate, i_spearman$p.value)
}

matched.corr$Spearman_rho <- as.numeric(matched.corr$Spearman_rho)
matched.corr$Spearman_pvalue <- as.numeric(matched.corr$Spearman_pvalue)

ggplot(matched.corr, aes(Marker, "PT/Met correlation\n(patient-matched)"))+
  geom_point(aes(color = Spearman_rho, size = -log(Spearman_pvalue)))+
    scale_colour_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-0.6, 0.6), na.value = "#b2182b")+
  scale_size(limits = c(0,13))+
  theme(panel.background = element_blank(), panel.border = element_rect(color = "black", fill = NA, size = 1), axis.text.x = element_text(angle = 90, hjust = 0), axis.title.y = element_blank(), legend.box = "horizontal")+
  scale_x_discrete(position = "top")
```

### CD4+ T cells

```{r CD4-marker-correlation}

sce_CD4 <- filterSCE(sce_immune_intra, clustername %in% c("T-CD4", "Treg"))
markers_include_CD4 <- c("CD38",  "CD279", "GITR", "CD7", "TCF7", "CXCL13", "FOXP3")

matched.corr <- as.data.frame(matrix(nrow = 0, ncol = 3))
colnames(matched.corr) <- c("Marker", "Spearman_rho", "Spearman_pvalue")

for (i in markers_include_CD4) {
  i_counts <- as.data.frame(assay(sce_CD4, "counts")[i,])
  colnames(i_counts) <- "i_counts"
  i_counts$patient.MetPT<- sce_CD4$patient.MetPT
  
  i_avg <- i_counts %>% group_by(patient.MetPT) %>% summarize(average = mean(i_counts))
  i_avg <- separate(i_avg, patient.MetPT, into = c("Patient", "Tissue.Class"), sep = "_")
  pat_incl <- table(i_avg$Patient)
  pat_incl <- names(pat_incl[pat_incl > 1])
  i_avg <- filter(i_avg, Patient %in% pat_incl)
  i_avg <- pivot_wider(i_avg, names_from = "Tissue.Class", values_from = "average")
  i_avg$metSite <- ROI_meta$Site.of.metastasis[match(i_avg$Patient, ROI_meta$Patient.ID)]
  i_spearman <- cor.test(i_avg$Met, i_avg$PT, method = "spearman")
  matched.corr[i,] <- c(i, i_spearman$estimate, i_spearman$p.value)
}

matched.corr$Spearman_rho <- as.numeric(matched.corr$Spearman_rho)
matched.corr$Spearman_pvalue <- as.numeric(matched.corr$Spearman_pvalue)

ggplot(matched.corr, aes(Marker, "PT/Met correlation\n(patient-matched)"))+
  geom_point(aes(color = Spearman_rho, size = -log(Spearman_pvalue)))+
    scale_colour_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-0.6, 0.6), na.value = "#b2182b")+
  scale_size(limits = c(0,13))+
  theme(panel.background = element_blank(), panel.border = element_rect(color = "black", fill = NA, size = 1), axis.text.x = element_text(angle = 90, hjust = 0), axis.title.y = element_blank(), legend.box = "horizontal")+
  scale_x_discrete(position = "top")

```

### Myeloid cells

```{r myeloid-marker-correlation}
markers_include_M <- c("CD38", "HLA-DR", "CLEC9A", "CD68", "CD163", "CD11b", "CD204", "GLUT1", "CD11c", "CD1c", "CD16", "MMP9", "CD274", "STING", "IDO", "CD33", "Arginase-1", "CD209", "CD40", "CD4", "CD14", "CD206", "LAMP3", "MPO", "CD303")

matched.corr <- as.data.frame(matrix(nrow = 0, ncol = 3))
colnames(matched.corr) <- c("Marker", "Spearman_rho", "Spearman_pvalue")

for (i in markers_include_M) {
  i_counts <- as.data.frame(assay(sce_M, "counts")[i,])
  colnames(i_counts) <- "i_counts"
  i_counts$patient.MetPT<- sce_M$patient.MetPT
  
  i_avg <- i_counts %>% group_by(patient.MetPT) %>% summarize(average = mean(i_counts))
  i_avg <- separate(i_avg, patient.MetPT, into = c("Patient", "Tissue.Class"), sep = "_")
  pat_incl <- table(i_avg$Patient)
  pat_incl <- names(pat_incl[pat_incl > 1])
  i_avg <- filter(i_avg, Patient %in% pat_incl)
  i_avg <- pivot_wider(i_avg, names_from = "Tissue.Class", values_from = "average")
  i_avg$metSite <- ROI_meta$Site.of.metastasis[match(i_avg$Patient, ROI_meta$Patient.ID)]
  i_spearman <- cor.test(i_avg$Met, i_avg$PT, method = "spearman")
  matched.corr[i,] <- c(i, i_spearman$estimate, i_spearman$p.value)
}

matched.corr$Spearman_rho <- as.numeric(matched.corr$Spearman_rho)
matched.corr$Spearman_pvalue <- as.numeric(matched.corr$Spearman_pvalue)

ggplot(matched.corr, aes(Marker, "PT/Met correlation\n(patient-matched)"))+
  geom_point(aes(color = Spearman_rho, size = -log(Spearman_pvalue)))+
    scale_colour_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-0.6, 0.6), na.value = "#b2182b")+
  scale_size(limits = c(0,13))+
  theme(panel.background = element_blank(), panel.border = element_rect(color = "black", fill = NA, size = 1), axis.text.x = element_text(angle = 90, hjust = 0), axis.title.y = element_blank(), legend.box = "horizontal")+
  scale_x_discrete(position = "top")

```

