---
title: "MyeloidPanel_spatial"
author: "SandraTietscher"
date: "2021-11-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This script looks at the spatial distribution of different myeloid cell phenotypes.


# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(dittoSeq)
library(cowplot)
library(scran)
library(edgeR)
library(CATALYST)
library(cytomapper)
library(vegan)
library(pairwiseAdonis)
library(tidyr)
library(dplyr)
library(ggpubr)
library(ggrepel)

# Read SingleCellExperiment object with subcluster myeloid cells
sce <- readRDS("output/SCEs/myeloid/myeloidSCE_M.rds")
```

# Compare phenotype proportions in intratumoral, stromal-enriched and margin regions of PT

```{r prepare-SCE-PT}
sce_PT <- filterSCE(sce, Tissue.Type == "PT")
sce_PT$patientLocation <- paste(sce_PT$Patient.ID, sce_PT$Location, sep = "_")
```

Use CATALYST's clrDR function to compute centered log-ratios on cluster proportions and visualize them in a lower-dimensional space (PCA).

```{r catalyst-clrDR-samples}
sce_PT$sample_id <- factor(sce_PT$patientLocation)
sce_PT$cluster_id <- sce_PT$M_cluster
sce_PT$condition <- factor(sce_PT$Location)
metadata(sce_PT)$cluster_codes <- data.frame(M_cluster = factor(sce_PT$M_cluster))

clrDR(sce_PT, dr = "PCA", by = "sample_id", k = "M_cluster", point_pal = metadata(sce)$colour_vectors$Location)
```

Permanova (Permutational Multivariate Analysis of Variance) on distance matrices to test whether the myeloid phenotypic composition differs significantly across different locations.

```{r permanova-location}

# Calculate Aitchisons distance (Euclidean distance on centered-log ratio)
patientLocation_tbl <- prop.table(table(sce_PT$patientLocation, sce_PT$cluster_id) + 0.00001, margin = 1)
patientLocation_clr <- compositions::clr(patientLocation_tbl)
dist <- vegdist(patientLocation_clr, method = "euclidean")

metadata <- data.frame(patient = sapply(strsplit(rownames(patientLocation_tbl), "_"), "[[", 1),
                       location = sapply(strsplit(rownames(patientLocation_tbl), "_"), "[[", 2))

# Check if dispersions between groups are different
mod_location <- betadisper(dist, metadata$location, type = "median", bias.adjust = FALSE,
       sqrt.dist = FALSE, add = FALSE)
permutest(mod_location, pairwise = TRUE, permutations = 999)

# Permanova test (overall and pairwise)
adonis(dist ~location,  data=metadata, permutations=9999)
pairwise.adonis(dist,metadata$location,p.adjust.m= 'BH')
```

Differential abundance by location (reference: margin regions)

```{r DA-location}
abundances <- table(sce_PT$M_cluster, sce_PT$patientLocation) 
abundances <- unclass(abundances) 

# Attach column metadata and relevel so that "margin" is the baseline group
meta <- colData(sce_PT)[match(colnames(abundances), sce_PT$patientLocation),]
location <- factor(meta$Location)
meta$location <- relevel(x=location, ref = "margin")
y.ab <- DGEList(abundances, samples=meta, group = meta$location)
design <- model.matrix(~Patient.ID + location, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Plot

## Stromal vs margin
  res <- glmQLFTest(fit.ab, coef="locationstromal")
  summary(decideTests(res))
  DA <- topTags(res, n = 20)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort by logFC
  order <- rownames(DA[order(DA$logFC),])
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("logFC: Stromal vs margin"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))

## Intratumoral vs margin
res <- glmQLFTest(fit.ab, coef="locationintratumoral")
  summary(decideTests(res))
  DA <- topTags(res, n = 20)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("logFC: Intratumoral vs margin"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

# Compare phenotype proportions in samples that are clinically classified as "hot" vs "cold"

```{r read-patho}
# Load csv with pathologist-annotated immune phenotype information
patho <- read.csv("data/immune_phenotypes_pathology_encoded.csv")
```

For PTs only - differential abundance testing using EdgeR.

```{r DA-hotcold-PT}
sce_PT$infiltration <- patho$Immune.phenotype.PT[match(sce_PT$Patient.ID, patho$Patient.ID)]

# Differential abundance testing
abundances <- table(sce_PT$cluster_id, sce_PT$Patient.ID) 
abundances <- unclass(abundances) 

# Attach column metadata and relevel so that "inflamed" is the baseline group
meta <- colData(sce_PT)[match(colnames(abundances), sce_PT$Patient.ID),]
infiltration <- factor(meta$infiltration)
meta$infiltration <- relevel(x=infiltration, ref = "inflamed")
y.ab <- DGEList(abundances, samples=meta, group = meta$infiltration)
design <- model.matrix(~infiltration, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance
res <- glmQLFTest(fit.ab)
summary(decideTests(res))
DA <- topTags(res, n= nrow(abundances))$table
DA$subtype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
DA <- DA[order(DA$logFC),]
DA$subtype <- factor(DA$subtype, levels = DA$subtype)

# Plot
ggplot(DA, aes(subtype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y="Excluded vs inflamed", x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For PTs and Met together - differential abundance testing using EdgeR.

```{r DA-hotcold-all}
sce$patient.site <- paste(sce$Patient.ID, ifelse(sce$Tissue.Type == "PT", "PT", "Met"), sep = "_")

# Clean up pathology table
patho1 <- patho[patho$Patient.ID != "ID17",] # Exclude the one patient whose metastatic sites have different annotations
patho_short <- patho1[,c(1,3,4)]
colnames(patho_short) <- c("Patient.ID", "PT", "Met")
patho_short <- pivot_longer(patho_short, 2:3, names_to = "site", values_to = "infiltration")
patho_short$patient.site <- paste(patho_short$Patient.ID, patho_short$site, sep = "_")

# Exclude patients not present in the pathology list
sce <- sce[,which(sce$Patient.ID %in% patho1$Patient.ID)]
sce$infiltration <- patho_short$infiltration[match(sce$patient.site, patho_short$patient.site)]

# Covert to binary characteristic (hot/cold)
sce$infiltration <- ifelse(sce$infiltration == "inflamed", "hot", "cold")

# Differential abundance testing
abundances <- table(sce$M_cluster, sce$patient.site) 
abundances <- unclass(abundances) 

# Attach column metadata and relevel so that "hot" is the baseline group
meta <- colData(sce)[match(colnames(abundances), sce$patient.site),]
infiltration <- factor(meta$infiltration)
meta$infiltration <- relevel(x=infiltration, ref = "hot")
y.ab <- DGEList(abundances, samples=meta, group = meta$infiltration)
design <- model.matrix(~infiltration, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance
res <- glmQLFTest(fit.ab)
summary(decideTests(res))
DA <- topTags(res, n= nrow(abundances))$table
DA$subtype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
DA <- DA[order(DA$logFC),]
DA$subtype <- factor(DA$subtype, levels = DA$subtype)

# Plot
ggplot(DA, aes(subtype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y="cold vs hot", x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

# Analyse spatial distribution of myeloid phenotypes in relation to tumor/stroma border

```{r border-distance-prep}
# Read in file with tumor/stroma mask information
tsmask <- read.csv("data/final_distanceTumorStroma_Cells_myeloid.csv")

# Read in file for bone images (was trained separately due to bad training quality)
tsmask_bone <- read.csv("data/myeloid_distanceTumorStroma_BONE_only_Cells.csv")
```

### Merge information about tumor/stroma border distance into main object

In some images, all cells have the value 65535 (max uint16 value). These are images where no tumor/stroma edge was identified, so no distances could be calculated. Exclude these for all mask-based analysis.

```{r merge-into-sce}
tsmask$ImageObjectNr <- paste(tsmask$ImageNumber, tsmask$ObjectNumber, sep="_")

# Match Object Number (via file name)
tsmask_bone$file <- paste0(tsmask_bone$Metadata_acname, "_a", tsmask_bone$Metadata_acid, "_ac_full.tiff")
tsmask_bone$ImageNumber <- sce$ImageNumber[match(tsmask_bone$file, sce$file)]
tsmask_bone$ImageObjectNr <- paste(tsmask_bone$ImageNumber, tsmask_bone$ObjectNumber, sep="_")
sce$ImageObjectNr <- paste(sce$ImageNumber, sce$CellNumber, sep="_")

# remove bone image cells from tsmask object
bone_img <- tsmask_bone$ImageObjectNr
tsmask <- filter(tsmask, !(ImageObjectNr %in% bone_img))

col_keep <- c("ImageObjectNr", "Intensity_MeanIntensity_TumourMask_Distance_B100", "Intensity_MinIntensityEdge_TumourMask_Distance_B100", "Intensity_MaxIntensityEdge_TumourMask_Distance_B100")
tsmask <- rbind(tsmask[,col_keep], tsmask_bone[,col_keep])


# Add mean, minimum and maximum border distances to object
sce$meanBorderDistance <- tsmask$Intensity_MeanIntensity_TumourMask_Distance_B100[match(sce$ImageObjectNr, tsmask$ImageObjectNr)]
sce$minBorderDistance <- tsmask$Intensity_MinIntensityEdge_TumourMask_Distance_B100[match(sce$ImageObjectNr, tsmask$ImageObjectNr)]
sce$maxBorderDistance <- tsmask$Intensity_MaxIntensityEdge_TumourMask_Distance_B100[match(sce$ImageObjectNr, tsmask$ImageObjectNr)]

# Set 65535 to NA
sce$meanBorderDistance <- ifelse(sce$meanBorderDistance == 65535, NA, sce$meanBorderDistance)
sce$minBorderDistance <- ifelse(sce$minBorderDistance == 65535, NA, sce$minBorderDistance)
sce$maxBorderDistance <- ifelse(sce$maxBorderDistance == 65535, NA, sce$maxBorderDistance)

# Assign binary mask status
sce$mask_bin <- ifelse(sce$meanBorderDistance > 0, "tumor", "stroma")

# Assign mask status where cells whose meanDistance is negative (outside tumor mask) but whose maxDistance is positive (inside tumor mask) are classified as "edge"
sce$mask <- ifelse(sce$meanBorderDistance > 0, "tumor", ifelse(sce$maxBorderDistance > 0, "edge", "stroma"))

# Subset cells (images) with no NAs
sce_sub <- sce[,which(!(is.na(sce$meanBorderDistance)))]
```

### Compare proportions of myeloid cells that are within a tumor mask in PT vs met
Intratumoral only!

```{r proportions-tumor-stroma}
sce_sub$patient_tissue <- paste(sce_sub$Patient.ID, sce_sub$Tissue.Type, sep = "_")

# Subset intratumoral
sce_intra <- sce_sub[,which(sce_sub$Location == "intratumoral")]
M_table <- as.data.frame(prop.table(table(sce_intra$patient_tissue, sce_intra$mask), margin = 1))
M_table <- separate(M_table, Var1, into = c("patient", "tissue"))

## Plot percentage that is within tumor mask
M_mask <- filter(M_table, Var2 == "tumor")
my_comparisons <- list( c("Brain", "PT"), c("Bone", "PT"), c("Liver", "PT"), c("SoftTissue", "PT"))
ggplot(M_mask, aes(tissue, Freq))+
  geom_boxplot()+
  ylab("Percentage of myeloid cells in tumor mask")+
  stat_compare_means(method = "anova")+
  stat_compare_means(comparisons = my_comparisons)
```

### Enrichment of myeloid cell subtypes within tumor vs stromal masks

Edge regions are classified as tumor regions for this analysis.

```{r Tcell-subtype-enrichment}
# Everything that is on the edge --> count into tumor mask
sce_sub$mask <- ifelse(sce_sub$mask == "stroma", "stroma", "tumor")

abundances <- table(sce_sub$M_cluster, sce_sub$mask)
logEnrichment_tumor <- as.data.frame(log2(abundances[,"tumor"]/abundances[,"stroma"]))
colnames(logEnrichment_tumor) <- "logEnrichment"
logEnrichment_tumor$M_cluster <- rownames(logEnrichment_tumor)

#colors
M_col <- c("#CC99FF", "#7F00FF", "#CC00CC", "#FF66FF", "#CC0066", "#9A6767", "#990000", 
           "#B30303", "#F03C3C", "#FF9999", "#FFB266", "#FF8000", "#CC6600", "#E6A73A",
           "#FFAE33", "#D2BD50", "#F6E176", "#F9D20E", "#FFEA33", "#D7FF33")
names(M_col) <- colnames(logEnrichment_tumor)

pos <- position_jitter(width = 0, height = 0.1, seed = 2)
ggplot(logEnrichment_tumor, aes(logEnrichment, "y")) +
  geom_jitter(aes(fill = M_cluster), position = pos, shape = 21, size = 3)+
  scale_fill_manual(values = M_col)+
  geom_label_repel(aes(label=M_cluster), size = 3, position = pos)+
  theme(axis.title.y= element_blank(), axis.text.y = element_blank(), 
        panel.border = element_rect(colour = "black", fill=NA, size=1), panel.background = element_blank())+
  geom_vline(xintercept = 0, linetype = "dashed")+
  xlab("Fold-enrichment in tumor mask [log2]")

```

### Correlation of marker expression with border distance

```{r marker-distance-curve}
# Re-scale arcsinh-transformed expression values (only over myeloid cells)
# assay(sce_sub, "scaled") <- t(scale(t(assay(sce_sub, "exprs"))))
# 
# markers_include <- c("CD38", "HLA-DR", "CLEC9A", "CD68", "CD163", "CD11b", "CD204", "CD11c", "CD1c", "CD16", "MMP9", "CD274", "STING", "IDO", "CD33", "Arginase-1", "CD209", "CD40", "CD4", "CD14", "CD206", "LAMP3", "MPO", "CD303", "GLUT1")
# 
# exprs <- t(assay(sce_sub, "scaled")[markers_include,])
# coldata <- as.data.frame(colData(sce_sub))
# coldata_ext <- cbind(coldata, exprs)
# coldata_ext <- pivot_longer(coldata_ext, cols = all_of(markers_include), values_to = "exprs", names_to = "marker")
# 
# ggplot(coldata_ext, aes(x = exprs, y = meanBorderDistance))+
#   #geom_point(color = "grey20", alpha = 0.1)+
#   #stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE)+
#   #scale_fill_distiller(palette=4, direction=1) +
#   facet_wrap(~marker, scales = "free_x", ncol = 7)+
#   coord_cartesian(ylim=c(-200, 50))+
#   geom_hline(yintercept = 0, linetype = "dashed")+
#   #stat_cor(method="spearman", label.x.npc = "middle", color = "black") + 
#   geom_smooth(method = "gam", color = "red", se=TRUE)+ 
#   xlab("Scaled arcsinh-transformed counts")+
#   theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
#         panel.background = element_blank(),
#         strip.background = element_blank())

```

```{r marker-distance-combined}
# markers_sub <- c("CD68", "CD163", "CD204", "CD11c", "CD1c", "MMP9", "CD274", "STING", "CD209", "CD14", "CD206", "LAMP3", "Arginase-1")
# 
# ggplot(filter(coldata_ext, marker %in% markers_sub), aes(x = exprs, y = meanBorderDistance, color = marker))+
#   coord_cartesian(ylim=c(-220, 50), xlim = c(-2,15))+
#   geom_hline(yintercept = 0, linetype = "dashed")+
#   geom_smooth(method = "loess", se=FALSE)+ 
#   scale_color_discrete()+
#   xlab("Scaled arcsinh-transformed counts")+
#   theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
#         panel.background = element_blank())

```




