---
title: "MyeloidPanel_subclustering"
author: "SandraTietscher"
date: "2021-11-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

This script performs separate subclustering of myeloid cells based on subtype and cell state markers.


# Load libraries and data

First, we will load the needed libraries and the `SingleCellExperiment` object, that was created in the previous section.

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(dittoSeq)
library(cowplot)
library(scran)
library(edgeR)
library(CATALYST)
library(cytomapper)

# Read SingleCellExperiment object
sce <- readRDS("output/SCEs/myeloid/myeloidSCE_clustered.rds")

# Subset myeloid cells
sce_M <- sce[,which(sce$celltype %in% c("myeloid"))]
sce_Mfull <- sce[,which(sce$celltype %in% c("myeloid","LAMP3_DC", "pDC", "neutrophil"))]
```

# Choose genes for clustering and UMAP

```{r choose-clustering-genes}

markers_include_Mfull <- c("CD38", "HLA-DR", "CLEC9A", "CD68", "CD163", "CD11b", "CD204", "CD11c", "CD1c", "CD16", "MMP9", "CD274", "STING", "IDO", "CD33", "Arginase-1", "CD209", "CD40", "CD4", "CD14", "CD206", "LAMP3", "CD303", "MPO")

markers_include_M <- c("CD38", "HLA-DR", "CLEC9A", "CD68", "CD163", "CD11b", "CD204", "CD11c", "CD1c", "CD16", "MMP9", "CD274", "STING", "IDO", "CD33", "Arginase-1", "CD209", "CD40", "CD4", "CD14", "CD206")
```

# UMAP representation

```{r compute-UMAP}

sce_M <- runUMAP(sce_M, exprs_values = "normalized", 
                     subset_row = rownames(assay(sce_M, "normalized")) %in% markers_include_M)

sce_Mfull <- runUMAP(sce_Mfull, exprs_values = "normalized", 
                     subset_row = rownames(assay(sce_M, "normalized")) %in% markers_include_Mfull)
```

## Visualize features and marker expression levels on UMAP

```{r visualize-features-umap, message = FALSE, fig.width=7, fig.height=5}
dittoDimPlot(sce_Mfull, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_Mfull)$colour_vectors$tissuetype, opacity = 0.3) 

dittoDimPlot(sce_Mfull, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                            color.panel = metadata(sce_Mfull)$colour_vectors$tissuetype, split.by = "Tissue.Type", opacity = 0.5)

dittoDimPlot(sce_Mfull, var = "Location", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_Mfull)$colour_vectors$Location, opacity = 0.3)

dittoDimPlot(sce_Mfull, var = "Location", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce_Mfull)$colour_vectors$Location, split.by = "Location",opacity = 0.5)

dittoDimPlot(sce_M, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                            color.panel = metadata(sce_Mfull)$colour_vectors$tissuetype, split.by = "Tissue.Type", opacity = 0.5)
```

```{r scaled-expression-umap, message=FALSE, fig.width=15, fig.height=10}
p.list <- list()
for(i in markers_include_Mfull){
  p.list[[i]] <- plotUMAP(sce_Mfull, colour_by = i, by_exprs_values = "scaled", 
                          point_size = 0.5, point_alpha = 1)  + scale_colour_gradient2(name = i, low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-3, 3), na.value = "#b2182b")
}

plot_grid(plotlist = p.list, ncol = 7)
```


# Run graph-based clustering

Run this only for the subset of myeloid cells that have not yet been assigned to a subtype (i.e. sce_M).

Use normalized counts for clustering, because this provides a better separation of noise and actual signal compared to arcsinh-transformed counts.  
Include only markers that were also used for UMAP.


```{r clustering}
cluster_counts_normalized <- subset(assay(sce_M, "normalized"), rownames(assay(sce_M, "normalized")) %in% markers_include_M)

g <- buildSNNGraph(cluster_counts_normalized, k=40, type = "jaccard")
clust<- igraph::cluster_louvain(g)$membership
table(clust)

# add cluster information to sce object
sce_M$cluster_k40<- as.factor(clust)

## Cluster annotation
sce_M$M_cluster <- paste0("M_c", sce_M$cluster_k40)

#plot on UMAP
dittoDimPlot(sce_M, var = "M_cluster", reduction.use = "UMAP", size = 0.5)
dittoDimPlot(sce_M, var = "M_cluster", reduction.use = "UMAP", size = 0.5, split.by = "M_cluster")
```

Include clusters into full myeloid object and all together on UMAP.

```{r merge-M-Mfull}
sce_Mfull$M_cluster <- sce_Mfull$clustername
M_ids <- colnames(sce_M)
sce_Mfull[,M_ids]$M_cluster <- sce_M$M_cluster

#select colors for subtypes
M_col <- c("#CC99FF", "#7F00FF", "#CC00CC", "#FF66FF", "#CC0066", "#9A6767", "#990000", 
           "#B30303", "#F03C3C", "#FF9999", "#FFB266", "#FF8000", "#CC6600", "#E6A73A",
           "#FFAE33", "#D2BD50", "#F6E176", "#F9D20E", "#FFEA33", "#D7FF33")
sce_Mfull$M_cluster <- factor(sce_Mfull$M_cluster)
names(M_col) <- levels(sce_Mfull$M_cluster)

dittoDimPlot(sce_Mfull, var = "M_cluster", reduction.use = "UMAP", size = 0.5, color.panel = M_col, do.label = TRUE, labels.size = 3)
dittoDimPlot(sce_Mfull, var = "M_cluster", reduction.use = "UMAP", size = 0.5, split.by = "M_cluster")

M_col_sub <- M_col[2:18]
dittoDimPlot(sce_M, var = "M_cluster", reduction.use = "UMAP", size = 0.5, color.panel = M_col_sub, do.label = TRUE, labels.size = 3)
dittoDimPlot(sce_M, var = "M_cluster", reduction.use = "UMAP", size = 0.5, color.panel = M_col_sub)
```

# Expression Heatmap of markers used for clustering/UMAP

```{r ExprHeatmap_type_state, fig.width = 11, fig.height=6}
sce_Mfull$sample_id <- sce_Mfull$ROI_ID
sce_Mfull$cluster_id <- sce_Mfull$M_cluster

rowData(sce_Mfull)$marker_class <- ifelse(rownames(sce_Mfull) %in% markers_include_Mfull, "type", "none")

metadata(sce_Mfull)$cluster_codes <- data.frame(M_cluster = factor(sce_Mfull$M_cluster))

plotExprHeatmap(sce_Mfull, 
features ="type", by = "cluster_id", k = "M_cluster", col_dend = TRUE, assay = "counts", scale = "last",
bars = TRUE, perc = TRUE, fun = "mean", k_pal = M_col)
```

# Plot subtype proportions per patient (separated by tissue type)

```{r proportions-per-patient}
sce_Mfull$sample_id <- sce_Mfull$Patient.ID
sce_Mfull$condition <- factor(sce_Mfull$Tissue.Type)
plotAbundances(sce_Mfull, k = "M_cluster", by = "sample_id", group_by = "condition")
```

# Save SCE objects

Also add subtype information to full object and save.

```{r save-SCE}
sce$M_cluster <- "none"
myeloid_ids <- colnames(sce_Mfull)
sce[,myeloid_ids]$M_cluster <- as.character(sce_Mfull$M_cluster)

saveRDS(sce, "output/SCEs/myeloid/myeloidSCE_subclustered.rds")
saveRDS(sce_M, "output/SCEs/myeloid/myeloidSCE_M_sub.rds")
saveRDS(sce_Mfull, "output/SCEs/myeloid/myeloidSCE_M.rds")
```

