---
title: "TcellPanel_03_spatial"
author: "SandraTietscher"
date: "2021-12-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(dittoSeq)
library(cowplot)
library(scran)
library(edgeR)
library(CATALYST)
library(cytomapper)
library(vegan)
library(pairwiseAdonis)
library(tidyr)
library(dplyr)
library(ggpubr)
library(ggrepel)
library(cytomapper)

# Read SingleCellExperiment object with subclustered T cells and subset T cells only
sce_full <- readRDS("output/SCEs/Tcell/TcellSCE_subclustered.rds")
sce <- sce_full[,which(sce_full$Tcell_subtype != "none")]
```

# Compare phenotype proportions in intratumoral, stromal-enriched and margin regions of PT

```{r prepare-SCE-PT}
sce_PT <- filterSCE(sce, Tissue.Type == "PT")
sce_PT$patientLocation <- paste(sce_PT$Patient.ID, sce_PT$Location, sep = "_")
```

Use CATALYST's clrDR function to compute centered log-ratios on cluster proportions and visualize them in a lower-dimensional space (PCA).

```{r catalyst-clrDR-samples}
sce_PT$sample_id <- factor(sce_PT$patientLocation)
sce_PT$cluster_id <- sce_PT$Tcell_subtype
sce_PT$condition <- factor(sce_PT$Location)
metadata(sce_PT)$cluster_codes <- data.frame(Tcell_subtype = factor(sce_PT$Tcell_subtype))

clrDR(sce_PT, dr = "PCA", by = "sample_id", k = "Tcell_subtype", point_pal = metadata(sce)$colour_vectors$Location)
```

Permanova (Permutational Multivariate Analysis of Variance) on distance matrices to test whether the centroids of the T cell phenotypic composition differs significantly across different locations. Aitchisons distance was chosen to calculate distance matrices

```{r permanova-location}

# Calculate Aitchisons distance (Euclidean distance on centered-log ratio)
patientLocation_tbl <- prop.table(table(sce_PT$patientLocation, sce_PT$cluster_id) + 0.00001, margin = 1)
patientLocation_clr <- compositions::clr(patientLocation_tbl)
dist <- vegdist(patientLocation_clr, method = "euclidean")

metadata <- data.frame(patient = sapply(strsplit(rownames(patientLocation_tbl), "_"), "[[", 1),
                       location = sapply(strsplit(rownames(patientLocation_tbl), "_"), "[[", 2))

# Check if dispersions between groups are different
mod_location <- betadisper(dist, metadata$location, type = "median", bias.adjust = FALSE,
       sqrt.dist = FALSE, add = FALSE)
permutest(mod_location, pairwise = TRUE, permutations = 999)

# Permanova test (overall and pairwise)
adonis(dist ~location,  data=metadata, permutations=9999)
pairwise.adonis(dist,metadata$location,p.adjust.m= 'BH')
```

Differential abundance by location (reference: margin regions)

```{r DA-location}
abundances <- table(sce_PT$Tcell_subtype, sce_PT$patientLocation) 
abundances <- unclass(abundances) 

# Attach column metadata and relevel so that "margin" is the baseline group
meta <- colData(sce_PT)[match(colnames(abundances), sce_PT$patientLocation),]
location <- factor(meta$Location)
meta$location <- relevel(x=location, ref = "margin")
y.ab <- DGEList(abundances, samples=meta, group = meta$location)
design <- model.matrix(~Patient.ID + location, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Plot

## Stromal vs margin
  res <- glmQLFTest(fit.ab, coef="locationstromal")
  summary(decideTests(res))
  DA <- topTags(res, n = 15)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort by logFC
  order <- rownames(DA[order(DA$logFC),])
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("logFC: Stromal vs margin"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))

## Intratumoral vs margin
res <- glmQLFTest(fit.ab, coef="locationintratumoral")
  summary(decideTests(res))
  DA <- topTags(res, n = 15)$table
  DA$celltype <- rownames(DA)
  DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
  
  # Sort same as above
  DA <- DA[order,]
  DA$celltype <- factor(DA$celltype, levels = DA$celltype)

ggplot(DA, aes(celltype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y=paste0("logFC: Intratumoral vs margin"), x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))

```


# Compare phenotype proportions in samples that are clinically classified as "hot" vs "cold"

```{r read-patho}
# Load csv with pathologist-annotated immune phenotype information
patho <- read.csv("data/immune_phenotypes_pathology_encoded.csv")
```

For PTs only - differential abundance testing using EdgeR.

```{r DA-hotcold-PT}
sce_PT$infiltration <- patho$Immune.phenotype.PT[match(sce_PT$Patient.ID, patho$Patient.ID)]

# Differential abundance testing
abundances <- table(sce_PT$Tcell_subtype, sce_PT$Patient.ID) 
abundances <- unclass(abundances) 

# Attach column metadata and relevel so that "inflamed" is the baseline group
meta <- colData(sce_PT)[match(colnames(abundances), sce_PT$Patient.ID),]
infiltration <- factor(meta$infiltration)
meta$infiltration <- relevel(x=infiltration, ref = "inflamed")
y.ab <- DGEList(abundances, samples=meta, group = meta$infiltration)
design <- model.matrix(~infiltration, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance
res <- glmQLFTest(fit.ab)
summary(decideTests(res))
DA <- topTags(res, n= nrow(abundances))$table
DA$subtype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
DA <- DA[order(DA$logFC),]
DA$subtype <- factor(DA$subtype, levels = DA$subtype)

# Plot
ggplot(DA, aes(subtype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y="Excluded vs inflamed", x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

For PTs and Met together - differential abundance testing using EdgeR.

```{r DA-hotcold-all}
sce$patient.site <- paste(sce$Patient.ID, ifelse(sce$Tissue.Type == "PT", "PT", "Met"), sep = "_")

# Clean up pathology table
patho1 <- patho[patho$Patient.ID != "ID17",] # Exclude the one patient whose metastatic sites have different annotations
patho_short <- patho1[,c(1,3,4)]
colnames(patho_short) <- c("Patient.ID", "PT", "Met")
patho_short <- pivot_longer(patho_short, 2:3, names_to = "site", values_to = "infiltration")
patho_short$patient.site <- paste(patho_short$Patient.ID, patho_short$site, sep = "_")

# Exclude patients not present in the pathology list
sce <- sce[,which(sce$Patient.ID %in% patho1$Patient.ID)]
sce$infiltration <- patho_short$infiltration[match(sce$patient.site, patho_short$patient.site)]

# Covert to binary characteristic (hot/cold)
sce$infiltration <- ifelse(sce$infiltration == "inflamed", "hot", "cold")

# Differential abundance testing
abundances <- table(sce$Tcell_subtype, sce$patient.site) 
abundances <- unclass(abundances) 

# Attach column metadata and relevel so that "hot" is the baseline group
meta <- colData(sce)[match(colnames(abundances), sce$patient.site),]
infiltration <- factor(meta$infiltration)
meta$infiltration <- relevel(x=infiltration, ref = "hot")
y.ab <- DGEList(abundances, samples=meta, group = meta$infiltration)
design <- model.matrix(~infiltration, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance
res <- glmQLFTest(fit.ab)
summary(decideTests(res))
DA <- topTags(res, n= nrow(abundances))$table
DA$subtype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
DA <- DA[order(DA$logFC),]
DA$subtype <- factor(DA$subtype, levels = DA$subtype)

# Plot
ggplot(DA, aes(subtype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y="cold vs hot", x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

Met only - differential abundance testing using EdgeR.

```{r DA-hotcold-met}
sce_Met <- sce[,which(sce$Tissue.Type != "PT")]

# Differential abundance testing
abundances <- table(sce_Met$Tcell_subtype, sce_Met$patient.site) 
abundances <- unclass(abundances) 

# Attach column metadata and relevel so that "hot" is the baseline group
meta <- colData(sce)[match(colnames(abundances), sce_Met$patient.site),]
infiltration <- factor(meta$infiltration)
meta$infiltration <- relevel(x=infiltration, ref = "hot")
y.ab <- DGEList(abundances, samples=meta, group = meta$infiltration)
design <- model.matrix(~infiltration, data = y.ab$samples)

# Estimate NB and QL dispersion
y.ab <- estimateDisp(y.ab, design, trend="none")
summary(y.ab$common.dispersion)
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
summary(fit.ab$var.prior)

# Test for differential abundance
res <- glmQLFTest(fit.ab)
summary(decideTests(res))
DA <- topTags(res, n= nrow(abundances))$table
DA$subtype <- rownames(DA)
DA$sign <- ifelse(DA$FDR<0.05, "yes", "no")
DA <- DA[order(DA$logFC),]
DA$subtype <- factor(DA$subtype, levels = DA$subtype)

# Plot
ggplot(DA, aes(subtype, y = logFC))+
  geom_col(aes(fill=sign))+
  scale_fill_manual(values = c("grey", "darkgreen"))+
  labs(y="cold vs hot", x="Cell type")+
  coord_flip()+
  labs(fill="Significant (FDR<0.05)")+
  theme(panel.background=element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1))
```

# Analyse spatial distribution of T cell phenotypes in relation to tumor/stroma border

```{r border-distance-prep}
# Read in file with tumor/stroma mask information
tsmask <- read.csv("data/final_distanceTumorStroma_Cells_Tcell.csv")

# Read in file for bone images (was trained separately due to bad training quality)
tsmask_bone <- read.csv("data/Tcell_distanceTumorStroma_BONE_only_Cells.csv")

```

### Merge information about tumor/stroma border distance into main object

In some images, all cells have the value 65535 (max uint16 value). These are images where no tumor/stroma edge was identified, so no distances could be calculated. Exclude these for all mask-based analysis.

```{r merge-into-sce}
tsmask$ImageObjectNr <- paste(tsmask$ImageNumber, tsmask$ObjectNumber, sep="_")

# Match Object Number (via file name)
tsmask_bone$file <- paste0(tsmask_bone$Metadata_acname, "_a", tsmask_bone$Metadata_acid, "_ac_full.tiff")
tsmask_bone$ImageNumber <- sce$ImageNumber[match(tsmask_bone$file, sce$file)]
tsmask_bone$ImageObjectNr <- paste(tsmask_bone$ImageNumber, tsmask_bone$ObjectNumber, sep="_")
sce$ImageObjectNr <- paste(sce$ImageNumber, sce$CellNumber, sep="_")

# remove bone image cells from tsmask object
bone_img <- tsmask_bone$ImageObjectNr
tsmask <- filter(tsmask, !(ImageObjectNr %in% bone_img))

col_keep <- c("ImageObjectNr", "Intensity_MeanIntensity_TumourMask_Distance_B100", "Intensity_MinIntensityEdge_TumourMask_Distance_B100", "Intensity_MaxIntensityEdge_TumourMask_Distance_B100")
tsmask <- rbind(tsmask[,col_keep], tsmask_bone[,col_keep])


# Add mean, minimum and maximum border distances to object
sce$meanBorderDistance <- tsmask$Intensity_MeanIntensity_TumourMask_Distance_B100[match(sce$ImageObjectNr, tsmask$ImageObjectNr)]
sce$minBorderDistance <- tsmask$Intensity_MinIntensityEdge_TumourMask_Distance_B100[match(sce$ImageObjectNr, tsmask$ImageObjectNr)]
sce$maxBorderDistance <- tsmask$Intensity_MaxIntensityEdge_TumourMask_Distance_B100[match(sce$ImageObjectNr, tsmask$ImageObjectNr)]

# Set 65535 to NA
sce$meanBorderDistance <- ifelse(sce$meanBorderDistance == 65535, NA, sce$meanBorderDistance)
sce$minBorderDistance <- ifelse(sce$minBorderDistance == 65535, NA, sce$minBorderDistance)
sce$maxBorderDistance <- ifelse(sce$maxBorderDistance == 65535, NA, sce$maxBorderDistance)

# Assign binary mask status
sce$mask_bin <- ifelse(sce$meanBorderDistance > 0, "tumor", "stroma")

# Assign mask status where cells whose meanDistance is negative (outside tumor mask) but whose maxDistance is positive (inside tumor mask) are classified as "edge"
sce$mask <- ifelse(sce$meanBorderDistance > 0, "tumor", ifelse(sce$maxBorderDistance > 0, "edge", "stroma"))

# Subset cells (images) with no NAs
sce_sub <- sce[,which(!(is.na(sce$meanBorderDistance)))]
```

Add mask information to full object (for cytomapper plotting)

```{r add-mask-full-object}
sce_full$ImageObjectNr <- paste(sce_full$ImageNumber, sce_full$CellNumber, sep="_")

# Add mean, minimum and maximum border distances to object
sce_full$meanBorderDistance <- tsmask$Intensity_MeanIntensity_TumourMask_Distance_B100[match(sce_full$ImageObjectNr, tsmask$ImageObjectNr)]
sce_full$minBorderDistance <- tsmask$Intensity_MinIntensityEdge_TumourMask_Distance_B100[match(sce_full$ImageObjectNr, tsmask$ImageObjectNr)]
sce_full$maxBorderDistance <- tsmask$Intensity_MaxIntensityEdge_TumourMask_Distance_B100[match(sce_full$ImageObjectNr, tsmask$ImageObjectNr)] 

# Set 65535 to NA
sce_full$meanBorderDistance <- ifelse(sce_full$meanBorderDistance == 65535, NA, sce_full$meanBorderDistance)
sce_full$minBorderDistance <- ifelse(sce_full$minBorderDistance == 65535, NA, sce_full$minBorderDistance)
sce_full$maxBorderDistance <- ifelse(sce_full$maxBorderDistance == 65535, NA, sce_full$maxBorderDistance)

# Assign binary mask status
sce_full$mask_bin <- ifelse(sce_full$meanBorderDistance > 0, "tumor", "stroma")

# Assign mask status where cells whose meanDistance is negative (outside tumor mask) but whose maxDistance is positive (inside tumor mask) are classified as "edge"
sce_full$mask <- ifelse(sce_full$meanBorderDistance > 0, "tumor", ifelse(sce_full$maxBorderDistance > 0, "edge", "stroma"))

# Subset cells (images) with no NAs
sce_sub_full <- sce_full[,which(!(is.na(sce_full$meanBorderDistance)))]
```


### Compare proportions of T cells that are within a tumor mask in PT vs met (intratumoral only)

```{r proportions-tumor-stroma}
sce_sub$patient_tissue <- paste(sce_sub$Patient.ID, sce_sub$Tissue.Type, sep = "_")

# Subset intratumoral
sce_intra <- sce_sub[,which(sce_sub$Location == "intratumoral")]
T_table <- as.data.frame(prop.table(table(sce_intra$patient_tissue, sce_intra$mask), margin = 1))
T_table <- separate(T_table, Var1, into = c("patient", "tissue"))

## Plot percentage that is within tumor mask
T_mask <- filter(T_table, Var2 == "tumor")
my_comparisons <- list( c("Brain", "PT"), c("Bone", "PT"), c("Liver", "PT"), c("SoftTissue", "PT"))
ggplot(T_mask, aes(tissue, Freq))+
  geom_boxplot()+
  ylab("Percentage of T cells cells in tumor mask")+
  stat_compare_means(method = "anova")+
  stat_compare_means(comparisons = my_comparisons)
```

CD8+ T cells only:

```{r CD8-proportions-tumor-stroma}
sce_intra_CD8 <- sce_intra[,which(sce_intra$clustername == "T_CD8")]

CD8_table <- as.data.frame(prop.table(table(sce_intra_CD8$patient_tissue, sce_intra_CD8$mask), margin = 1))
CD8_table <- separate(CD8_table, Var1, into = c("patient", "tissue"))

## Plot percentage that is within tumor mask
CD8_mask <- filter(CD8_table, Var2 == "tumor")
my_comparisons <- list( c("Brain", "PT"), c("Bone", "PT"), c("Liver", "PT"), c("SoftTissue", "PT"))
ggplot(CD8_mask, aes(tissue, Freq))+
  geom_boxplot()+
  ylab("Percentage of T cells cells in tumor mask")+
  stat_compare_means(method = "anova")+
  stat_compare_means(comparisons = my_comparisons)

```


### Compare proportions of T cells that are within a tumor mask in hot vs cold

```{r proportions-tumor-stroma-hotcold}
T_table <- as.data.frame(prop.table(table(sce_sub$patient_tissue, sce_sub$mask), margin = 1))
T_table$infiltration <- sce_sub$infiltration[match(T_table$Var1, sce_sub$patient_tissue)]
T_table <- separate(T_table, Var1, into = c("patient", "tissue"))

## Plot percentage that is within tumor mask
T_mask <- filter(T_table, Var2 == "tumor")
ggplot(T_mask, aes(infiltration, Freq))+
  geom_boxplot()+
  ylab("Percentage of T cells in tumor mask")+
  stat_compare_means(method = "wilcox")


# Subset intratumoral
sce_intra <- sce_sub[,which(sce_sub$Location == "intratumoral")]
T_table <- as.data.frame(prop.table(table(sce_intra$patient_tissue, sce_intra$mask), margin = 1))
T_table$infiltration <- sce_intra$infiltration[match(T_table$Var1, sce_intra$patient_tissue)]
T_table <- separate(T_table, Var1, into = c("patient", "tissue"))

    ## Only for PT
    T_mask <- filter(T_table, Var2 == "tumor" & tissue == "PT")
    ggplot(T_mask, aes(infiltration, Freq))+
      geom_boxplot()+
      ylab("Percentage of T cells cells in tumor mask")+
      stat_compare_means(method = "wilcox")

    ## Only for met
    T_mask <- filter(T_table, Var2 == "tumor" & tissue != "PT")
    ggplot(T_mask, aes(infiltration, Freq))+
      geom_boxplot()+
      ylab("Percentage of T cells cells in tumor mask")+
      stat_compare_means(method = "wilcox")
```

### Enrichment of T cell subtypes within tumor vs stromal masks

Edge regions are classified as tumor regions for this analysis.

```{r Tcell-subtype-enrichment}
# Everything that is on the edge --> count into tumor mask
sce_sub$mask <- ifelse(sce_sub$mask == "stroma", "stroma", "tumor")

abundances <- table(sce_sub$Tcell_subtype, sce_sub$mask)
logEnrichment_tumor <- as.data.frame(log2(abundances[,"tumor"]/abundances[,"stroma"]))
colnames(logEnrichment_tumor) <- "logEnrichment"
logEnrichment_tumor$Tcell_subtype <- rownames(logEnrichment_tumor)

#colors
T_col <- c("#666600", "#003319", "#CCCC00", "#00CC00", "#00FF00", "#B2FF66", "#CCFFCC", "#66FFFF", "#00CCCC", "#3333FF", "#99CCFF")
names(T_col) <- rownames(abundances)

pos <- position_jitter(width = 0, height = 0.1, seed = 2)
ggplot(logEnrichment_tumor, aes(logEnrichment, "y")) +
  geom_jitter(aes(fill = Tcell_subtype), position = pos, shape = 21, size = 3)+
  scale_fill_manual(values = T_col)+
  geom_label_repel(aes(label=Tcell_subtype), size = 3, position = pos)+
  theme(axis.title.y= element_blank(), axis.text.y = element_blank(), 
        panel.border = element_rect(colour = "black", fill=NA, size=1), panel.background = element_blank())+
  geom_vline(xintercept = 0, linetype = "dashed")+
  xlab("Fold-enrichment in tumor mask [log2]")

```

PT only.

```{r mask-subtype-enrichment-PT}
sce_sub_PT <- sce_sub[,which(sce_sub$Tissue.Type == "PT")]

abundances <- table(sce_sub_PT$Tcell_subtype, sce_sub_PT$mask)
logEnrichment_tumor <- as.data.frame(log2(abundances[,"tumor"]/abundances[,"stroma"]))
colnames(logEnrichment_tumor) <- "logEnrichment"
logEnrichment_tumor$Tcell_subtype <- rownames(logEnrichment_tumor)

ggplot(logEnrichment_tumor, aes(logEnrichment, "y")) +
  geom_jitter(aes(fill = Tcell_subtype), position = pos, shape = 21, size = 3)+
  scale_fill_manual(values = T_col)+
  geom_text_repel(aes(label=Tcell_subtype), size = 3, position = pos)+
  theme(axis.title.y= element_blank(), axis.text.y = element_blank(), 
        panel.border = element_rect(colour = "black", fill=NA, size=1), panel.background = element_blank())+
  geom_vline(xintercept = 0, linetype = "dashed")+
  xlab("Fold-enrichment in tumor mask [log2]")
```

Met only.

```{r mask-subtype-enrichment-met}
sce_sub_met <- sce_sub[,which(sce_sub$Tissue.Type != "PT")]

abundances <- table(sce_sub_met$Tcell_subtype, sce_sub_met$mask)
logEnrichment_tumor <- as.data.frame(log2(abundances[,"tumor"]/abundances[,"stroma"]))
colnames(logEnrichment_tumor) <- "logEnrichment"
logEnrichment_tumor$Tcell_subtype <- rownames(logEnrichment_tumor)

ggplot(logEnrichment_tumor, aes(logEnrichment, "y")) +
  geom_jitter(aes(fill = Tcell_subtype), position = pos, shape = 21, size = 3)+
  scale_fill_manual(values = T_col)+
  geom_text_repel(aes(label=Tcell_subtype), size = 3, position = pos)+
  theme(axis.title.y= element_blank(), axis.text.y = element_blank(), 
        panel.border = element_rect(colour = "black", fill=NA, size=1), panel.background = element_blank())+
  geom_vline(xintercept = 0, linetype = "dashed")+
  xlab("Fold-enrichment in tumor mask [log2]")
```

# Cytomapper plotting

```{r}
path.to.images <- "data/example_Tcell_masks"
all_masks <- loadImages(path.to.images, pattern = "_mask.tiff")
# Add image number
#imagenr <- read.csv("data/masks/ProteinPanel_masks/image_number_HIER.csv", header=FALSE)[1]
mcols(all_masks)$ROI_ID <- "TMA266_ROI_f01_immune"
# Scale images
all_masks <- scaleImages(all_masks, 2^16-1)

cur_img <- getImages(all_masks, "20200617_ZTMA266_immune_LKST_row1_to_row7_s0_a6_ac_ilastik_s2_Probabilities_mask")

color <- c("grey", "red", rep("yellow",10))
#names(color) <- unique(sce_full$celltype)

# Color by cell type
plotCells(cur_img, object = sce_full,
            img_id = "ROI_ID", cell_id = "CellNumber",
            colour_by = "celltype",
            colour = list(celltype = c(epithelial = "#00FFFF",
                          stromal_ne = "#3333FF",
                          plasma_cell = "#009900",
                          endothelial = "#FF9933",
                          myeloid = "#FF0000",
                          T_cell = "#00FF00",
                          NK_cell = "#00FF00",
                          undefined = "#C0C0C0",
                          `HLA-DR_high` = "#C0C0C0",
                          B_cell = "#FF3399",
                          neutrophil = "#FF0000")),
            scale_bar = NULL,
            image_title = NULL)


# Color by mask
plotCells(cur_img, object = sce_full,
            img_id = "ROI_ID", cell_id = "CellNumber",
            colour_by = "mask",
                      colour = list(celltype = c(epithelial = "#00FFFF",
                          stromal_ne = "#3333FF",
                          plasma_cell = "#009900",
                          endothelial = "#FF9933",
                          myeloid = "#FF0000",
                          T_cell = "#00FF00",
                          NK_cell = "#00FF00",
                          undefined = "#C0C0C0",
                          `HLA-DR_high` = "#C0C0C0",
                          B_cell = "#FF3399",
                          neutrophil = "#FF0000")),
            scale_bar = NULL,
            image_title = NULL)
          
```
