---
title: "TcellPanel_QC_1stClustering"
author: "SandraTietscher"
date: "2021-11-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

In this script, we work on the inital T cell-panel SCE object to
- Add different count transformations needed for subsequent analysis steps 
- Perform basic QC and filtering steps (cell-level and image-level filtering)
- Run UMAP and perform a first clustering step on the full dataset

## Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(scran)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(cowplot)
library(dittoSeq)
library(tidyverse)
library(CATALYST)

# Load SingleCellExperiment object
sce <- readRDS("output/SCEs/Tcell/TcellSCE_initial.rds")
```

# Data transformation, QC and cell-level/image-level filtering

## Generate colour vectors for main features

Define colours for 
1. the tissue type
2. the location of the punch (intratumoral, margin, stromal)
3. the molecular subtype as defined in the clinic

```{r generate-colour-vectors}
colour_vectors <- list()

col_vec_tissuetype <- c("#CCCC00", "#4C9900", "#FF8000", "#FF3333", "#AC16AC")
names(col_vec_tissuetype) <- c("SoftTissue", "PT", "Bone", "Liver", "Brain")

col_vec_location <- c("#00FFFF","#FF00FF", "#004C99")
names(col_vec_location) <- c("intratumoral", "stromal", "margin")

col_vec_subtype <- c("#8900FF", "#0022FF", "#00B3FF", "#0080FF")
names(col_vec_subtype) <- unique(sce$molecular.subtype)

colour_vectors$tissuetype <- col_vec_tissuetype
colour_vectors$Location <- col_vec_location
colour_vectors$subtype <- col_vec_subtype

# Save in metadata slot
metadata(sce)$colour_vectors <- colour_vectors
```

## Transform the data

Different data transformations are stored in different assay slots:  
arcsinh transformation -> `exprs` slot  
scaled arcsinh transformation -> `scaled` slot  
99th percentile normalization -> `normalized` slot  

```{r data-transformation}
assay(sce, "exprs") <- asinh(counts(sce))
assay(sce, "scaled") <- t(scale(t(assay(sce, "exprs"))))

norm_counts <- t(apply(assay(sce, "counts"), 1, function(x)(x-min(x))/(quantile(x, 0.99)-min(x))))
norm_counts <- t(apply(norm_counts, 1, function(x) pmin(x, 1)))
assay(sce, "normalized", withDimnames = FALSE) <- norm_counts
```

## Remove images with less than 100 cells

Cell type proportion/phenotype analysis on this images will likely not be meaningful.

```{r cellnumber-filter}
cellnumber_image <- as.data.frame(table(sce$ROI_ID))

# define images with less than 100 cells ###
imageIDs_rm <- filter(cellnumber_image, Freq < 100)[,1]

# Look at the image metadata before removing them
sce.small <- sce[,which(sce$ROI_ID %in% imageIDs_rm)]
table(sce.small$Tissue.Type)
table(sce.small$Location)

# remove
sce <- sce[,which(!(sce$ROI_ID %in% imageIDs_rm))]
```

## Remove very small and very large cells

Cells smaller than 7 pixels and larger than 600 pixels are likely to be a segmentation artefact.

```{r cell-size-filter, fig.width=10}
# Count and remove cells smaller than 7 pixels
sum(sce$Area < 7)
sce$include <- sce$Area >= 7
sce <- sce[, which(sce$include == TRUE)]

# Count and remove cells larger than 600 pixels
sum(sce$Area > 600)
sce$include <- sce$Area <= 600
sce <- sce[, which(sce$include == TRUE)]

# Plot average cell size per tissue type
colData <- as.data.frame(colData(sce))
ggplot(colData, aes(x = Tissue.Type, y = Area)) + 
  geom_violin()+
  geom_boxplot(outlier.size = 0.5, width = 0.2)+
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## QC: Mean marker intensities 

Visualization of the mean marker intensites per ROI, using the `aggregateAcrossCells` function of the `scater` package, which returns a `SingleCellExperiment` object for further downstream processing. 

```{r mean-intensity-heatmap, fig.height=10, fig.width = 10}
mean_sce <- aggregateAcrossCells(sce, ids = sce$ROI_ID, statistics = "mean")
assay(mean_sce, "arcsinh") <- asinh(assay(mean_sce, "counts")) 

# Centered and scaled
dittoHeatmap(mean_sce, assay = "arcsinh",
            annot.by = c("Tissue.Type", "Location"),
            cluster_cols = TRUE, heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            annotation_colors = list(Tissue.Type = metadata(sce)$colour_vectors$tissuetype,
                                     Location = metadata(sce)$colour_vectors$Location),
            breaks = seq(-3, 3, length.out = 101))

# Without DNA
markers_incl <- rownames(mean_sce)[!(rownames(mean_sce) %in% c("DNA1", "DNA2"))]
dittoHeatmap(mean_sce[markers_incl,], assay = "arcsinh",
            annot.by = c("Tissue.Type", "Location"),
            cluster_cols = TRUE, heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            annotation_colors = list(Tissue.Type = metadata(sce)$colour_vectors$tissuetype,
                                     Location = metadata(sce)$colour_vectors$Location),
            breaks = seq(-3, 3, length.out = 101))
```

## QC: Low-dimensional representation of image-level information

Use PCA to compute low-dimensional representations of image-level median intensity counts in order to check for potential batch effects.

```{r PCA-image-level, message = FALSE}
# Principal component analysis
mean_sce <- runPCA(mean_sce, exprs_values = "arcsinh", ncomponents = 10, 
                   subset_row = !(rownames(mean_sce) %in% c("DNA1", "DNA2", "HH3")))
plotPCA(mean_sce, colour_by = "Tissue.Type", shape_by = "Location", 
        point_alpha = 1, point_size = 4) + 
  scale_colour_manual(values = metadata(sce)$colour_vectors$tissuetype)

# Color by detector change status (pre/post)
plotPCA(mean_sce, colour_by = "DetectorChange", shape_by = "Location", 
        point_alpha = 1, point_size = 4)

```

# First-level clustering (full dataset)

### Choose genes for clustering and UMAP

Choose only cell type markers that should be used for the first clustering step as well as UMAP representation. Exclude cell state markers or cell subtype markers that will only be needed for subclustering.

```{r choose-clustering-genes}
markers_UMAP <- c("MPO", "SMA", "panCK", "CD38", "HLA-DR", "CD15", "CD68", "CD163", "CD20", "CD3", "CD11c", "CD7","FOXP3", "CD8a", "CD140b", "Caveolin-1", "CD4", "CD31_vWF", "E/P-Cadherin")
```

## UMAP representation

For UMAP calculation, the noramlized counts are used. As the goal here is only to identify main cell types, only the main cell type markers are included. To speed up UMAP calculation and plotting, only a subset of cells is used.


### Calculate UMAP on a subset of the cells

```{r compute-UMAP}
sce$cellID <- rownames(colData(sce))

# For images with less than 300 cells: Use all cells
as.data.frame(table(sce$ImageNumber)) %>% 
  filter(Freq < 300) %>% 
  pull(Var1) -> small.images
cell_idx_smallImages <- sce[, which(sce$ImageNumber %in% small.images)]$cellID

# For image with more than 300 cells: Randomly sample 300 cells
colData(sce[,which(!(sce$ImageNumber %in% small.images))]) %>%
  as.data.frame() %>%
  group_by(ImageNumber) %>%
  sample_n(300) %>% 
  pull(cellID) -> cell_idx_largeImages

cell_idx <- c(cell_idx_largeImages, cell_idx_smallImages)
sce_sub <- sce[,cell_idx]

# Run Dimensional reduction
sce_sub <- runUMAP(sce_sub, exprs_values = "normalized", 
                     subset_row = rownames(assay(sce_sub, "normalized")) %in% markers_UMAP)
```

### Visualize features on UMAP

```{r visualize-features-umap, message = FALSE, fig.width=6, fig.height=5}

dittoDimPlot(sce_sub, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$tissuetype, opacity = 0.3) 

dittoDimPlot(sce_sub, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$tissuetype, split.by = "Tissue.Type", opacity = 0.5) 

dittoDimPlot(sce_sub, var = "Location", reduction.use = "UMAP", size = 0.5,  color.panel = metadata(sce)$colour_vectors$Location, opacity = 0.3)

dittoDimPlot(sce_sub, var = "Location", reduction.use = "UMAP", size = 0.5, color.panel = metadata(sce)$colour_vectors$Location, split.by = "Location",opacity = 0.5)

```

### Visualize expression on UMAP

Using scaled arcsinh-transformed counts.

```{r scaled-expression-umap, message=FALSE, fig.width=35, fig.height=25}
p.list <- list()
for(i in rownames(sce_sub)){
  p.list[[i]] <- plotUMAP(sce_sub, colour_by = i, by_exprs_values = "scaled", 
                          point_size = 0.5, point_alpha = 1)  + scale_colour_gradient2(name = i, low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-3, 3), na.value = "#b2182b")
}

plot_grid(plotlist = p.list, ncol = 7)
```

## Run graph-based clustering

Use normalized counts for clustering, because this provides a better separation of noise and actual signal compared to arcsinh-transformed counts.  
Include only markers that are relevant for the general distinction between tumor and stromal cells

```{r clustering}
cluster_counts_normalized <- subset(assay(sce, "normalized"), rownames(assay(sce, "normalized")) %in% markers_UMAP)

g <- buildSNNGraph(cluster_counts_normalized, k=5, type = "jaccard")
clust<- igraph::cluster_louvain(g)$membership
table(clust)

# add cluster information to sce object
sce$cluster_k5 <- as.factor(clust)

# add cluster information to sce_sub object
sub_cellnames <- colnames(sce_sub)
cluster_sub <- sce[,sub_cellnames]$cluster_k5
sce_sub$cluster_k5 <- cluster_sub

# remove tiny cluster 20
sce <- sce[,which(sce$cluster_k5 != 20)]
sce_sub <- sce_sub[,which(sce_sub$cluster_k5 != 20)]
```


### Cluster annotation

Annotate each cluster with three metadata columns:
`class` is either epithelial, stromal, immune or undefined
`celltype` is a more precise celltype annotation
`clustername` is a unique name composed of celltype label and a number

```{r cluster_annotation, fig.width=7, fig.height=5}
labels <- read.csv("data/cluster_annotations/Tcell_full_cluster_annotation.csv")

label_class <- labels$class[match(sce$cluster_k5, labels$cluster)]
sce$cell_class <- label_class

label_celltype <- labels$celltype[match(sce$cluster_k5, labels$cluster)]
sce$celltype <- label_celltype

label_clustername <- labels$clustername[match(sce$cluster_k5, labels$cluster)]
sce$clustername <- label_clustername

#plot on UMAP
label_class_sub <- labels$class[match(sce_sub$cluster_k5, labels$cluster)]
sce_sub$cell_class <- label_class_sub
label_celltype_sub <- labels$celltype[match(sce_sub$cluster_k5, labels$cluster)]
sce_sub$celltype <- label_celltype_sub
label_clustername_sub <- labels$clustername[match(sce_sub$cluster_k5, labels$cluster)]
sce_sub$clustername <- label_clustername_sub

dittoDimPlot(sce_sub, var = "cell_class", reduction.use = "UMAP", size = 0.5)
dittoDimPlot(sce_sub, var = "celltype", reduction.use = "UMAP", size = 0.5)
dittoDimPlot(sce_sub, var = "clustername", reduction.use = "UMAP", size = 0.5)
```

### Plot expression heatmap for ALL markers by cluster name

```{r expression-heatmap-all, fig.width=10, fig.height=6}
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% markers_UMAP, "type", "state")
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% c("DNA1", "DNA2"), "none", rowData(sce)$marker_class)

sce$sample_id <- sce$ROI_ID
sce$cluster_id <- sce$cluster_k5
metadata(sce)$cluster_codes <- data.frame(custom = factor(sce$cluster_k5), clustername = factor(sce$clustername))

plotMultiHeatmap(sce, 
hm1 = "type", hm2 = "state", 
k = "clustername", col_dend = c(TRUE, TRUE), assay = "counts", scale = "last", bars = TRUE, perc = TRUE, fun = "mean")
```

# Save SCE object

```{r save-sce}
saveRDS(sce, "output/SCEs/Tcell/TcellSCE_clustered.rds")
#saveRDS(sce_sub, "output/SCEs/Tcell/TcellSCE_clustered_sub_umap.rds")
```
