---
title: "TumorPanel_01_QC_1stClustering"
author: "LauraKuett"
date: "2021-15-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This script visualizes image level intensities for tumor panel and performs first clustering step to separated out epithelial cells for downstream analysis. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and data

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(ggridges)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(cowplot)
library(dittoSeq)
library(scales)
library(tidyverse)
library(ggpubr)
library(scran)
library(CATALYST)

# Read SingleCellExperiment object
sce <- readRDS("output/SCEs/tumor/tumorSCE_initial.rds")
```

# Remove control ROIs

Placenta ROIs were acquired as an antibody control, but should not be included in the analysis.

```{r remove-control-images}
sce <- sce[, which(!(sce$Tissue.Type %in% c("ROI_yy", "ROI_xx", "ROI_zz")))]
```

# Remove empty channel

No antibody in Nd144 -> remove.

```{r remove-empty-channel}
sce <- sce[rownames(sce) != "empty",]
```


# Generate colour vectors for main features

Primarily, we will define colours for 
1. the tissue type
2. the location of the punch (intratumoral, margin, stromal)
3. the patient ID

```{r generate-colour-vectors}
colour_vectors <- list()

col_vec_tissuetype <- c( "#4C9900", "#FF8000", "#AC16AC", "#FF3333","#CCCC00")
names(col_vec_tissuetype) <- c( "PT", "Bone", "Brain","Liver","SoftTissue")

col_vec_location <- c("#3399FF","#9B4AF7", "#3333FF")
names(col_vec_location) <- c("intratumoral", "stromal", "margin")

col_vec_PatientID <- dittoColors(reps = 1)[seq_len(length(unique(sce$Patient.ID)))]
names(col_vec_PatientID ) <- unique(sce$Patient.ID)
col_vec_molecular_subtype <- c("#FC4E07","#E7B800","#00AFBB","gray")
names(col_vec_molecular_subtype) <- c("Her2","luminal A", "luminal B","TN")

colour_vectors$tissuetype <- col_vec_tissuetype
colour_vectors$Location <- col_vec_location
colour_vectors$PatientID <- col_vec_PatientID

# Save in metadata slot
metadata(sce)$colour_vectors <- colour_vectors

```

# Transform the data

Different data transformations are stored in different assay slots:  
arcsinh transformation -> `exprs` slot  
scaled arcsinh transformation -> `scaled` slot (per column mean centered and divided by standard deviation)
simple 0-1 normalization (using 99th percentile as 1) -> `normalized` slot  

```{r data-transformation}
assay(sce, "exprs") <- asinh(counts(sce))
assay(sce, "scaled") <- t(scale(t(assay(sce, "exprs"))))

norm_counts <- t(apply(assay(sce, "counts"), 1, function(x)(x-min(x))/(quantile(x, 0.99)-min(x))))
norm_counts <- t(apply(norm_counts, 1, function(x) pmin(x, 1)))
assay(sce, "normalized", withDimnames = FALSE) <- norm_counts
```

# Visualize image-level information


## Number of cells

First, we will visualize the number of cells across various cell-level metadata entries.
Images with less than 100 cells are removed.


### 2. Number of cells per image

```{r no-cells-image, fig.width=6, fig.height=6}
cellnumber_image <- as.data.frame(table(sce$ROI_ID))

# lowest numbers
sort(cellnumber_image$Freq)[1:30]

# define images with less than 100 cells ###
imageIDs_rm <- filter(cellnumber_image, Freq < 100)[,1]

# Look at the image metadata before removing them
sce.small <- sce[,which(sce$ROI_ID %in% imageIDs_rm)]
table(sce.small$Tissue.Type)
table(sce.small$Location)

# remove
sce <- sce[,which(!(sce$ROI_ID %in% imageIDs_rm))]
```


# Remove cells that are too small or too big


```{r cell-size, fig.width=10}
# Count and remove cells smaller than 7 pixels
sum(sce$Area < 7)
sce$include <- sce$Area >= 7
sce <- sce[, which(sce$include == TRUE)]

# Count and remove cells larger than 600 pixels
sum(sce$Area > 600)
sce$include <- sce$Area <= 600
sce <- sce[, which(sce$include == TRUE)]

```


## Mean marker intensities 

Next step: Visualization of the mean marker intensities per ROI.
For this, we will use the `aggregateAcrossCells` function of the `scater` package, which returns a `SingleCellExperiment` object for further downstream processing. 
We will arcsinh transform the mean intensity values prior to plotting.

```{r mean-intensity-heatmap, fig.height=10, fig.width = 10}
mean_sce <- aggregateAcrossCells(sce, ids = sce$ROI_ID, statistics = "mean")
assay(mean_sce, "arcsinh") <- asinh(assay(mean_sce, "counts")) 


# Centered and scaled
dittoHeatmap(mean_sce, assay = "arcsinh",
            annot.by = c("Tissue.Type", "Location"),
            cluster_cols = TRUE, heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            annotation_colors = list(Tissue.Type = metadata(sce)$colour_vectors$tissuetype,
                                     Location = metadata(sce)$colour_vectors$Location),
            breaks = seq(-3, 3, length.out = 101))
```

## Low-dimensional representation of image-level information

Next, we can use the summarized `SingleCellExperiment` object to compute low-dimensional representations of image-level median intensity counts. For this, we will use a PCA representation of the images.

```{r PCA-image-level, message = FALSE}
# Principal component analysis
mean_sce <- runPCA(mean_sce, exprs_values = "arcsinh", ncomponents = 10, 
                   subset_row = !(rownames(mean_sce) %in% c("DNA1", "DNA2", "HH3")))
plotPCA(mean_sce, colour_by = "Tissue.Type", shape_by = "Location", 
        point_alpha = 1, point_size = 4) + 
  scale_colour_manual(values = metadata(sce)$colour_vectors$tissuetype)

```


## Average cell size per Tissue Type

```{r cell-size-tissuetype, fig.width=7, fig.height=5}
colData <- as.data.frame(colData(sce))
colData$Tissue.Type <- factor(colData$Tissue.Type)
colData$Tissue.Type <- relevel(colData$Tissue.Type, "PT")

ggplot(colData, aes(x = Tissue.Type, y = Area)) + 
  geom_violin()+
  geom_boxplot(outlier.size = 0.5, width = 0.2)+
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```


# UMAP representation of all cells

For UMAP calculation, the normalized counts are used for a subset of cells to speed up the performance.


## Calculate UMAP on a subset of the cells using markers that are also used for downstream clustering to seperated out epithelial cells

```{r compute-UMAP}

markers_include <- c("SMA","panCK","Vimentin","CK8_18_19","CD68", "CK14", "CD20","CD3", "CD11c",
                     "HER2", "CK5","GATA3","CK7","CD8a",
                     "Fibronectin","CD4","CD31_vWF","E/P-Cadherin")


sce$cellID <- rownames(colData(sce))

# For images with less than 300 cells: Use all cells
as.data.frame(table(sce$ImageNumber)) %>% 
  filter(Freq < 300) %>% 
  pull(Var1) -> small.images
cell_idx_smallImages <- sce[, which(sce$ImageNumber %in% small.images)]$cellID

# For image with more than 300 cells: Randomly sample 300 cells
colData(sce[,which(!(sce$ImageNumber %in% small.images))]) %>%
  as.data.frame() %>%
  group_by(ImageNumber) %>%
  sample_n(300) %>% 
  pull(cellID) -> cell_idx_largeImages

cell_idx <- c(cell_idx_largeImages, cell_idx_smallImages)
sce_sub <- sce[,cell_idx]



# Run Dimensional reduction
sce_sub <- runUMAP(sce_sub,n_neighbors =15,  exprs_values = "normalized", 
                     subset_row = rownames(assay(sce_sub, "normalized")) %in% markers_include)


```
## Visualize features on UMAP

```{r visualize-features-umap, message = FALSE, fig.width=10, fig.height=8}

dittoDimPlot(sce_sub, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$tissuetype, opacity = 0.3) 

dittoDimPlot(sce_sub, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$tissuetype, split.by = "Tissue.Type", opacity = 0.5) 

dittoDimPlot(sce_sub, var = "Location", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$Location, opacity = 0.3)

dittoDimPlot(sce_sub, var = "Location", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$Location, split.by = "Location",opacity = 0.5)

dittoDimPlot(sce_sub, var = "molecular.subtype", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$Location, split.by = "Location",opacity = 0.5)
```

## Visualize expression on UMAP

Visualize the expression of each marker.

### 1. Arcsinh-transformed counts

```{r arcsinh-expression-umap, fig.width=35, fig.height=25}
#ER not working for plotting on UMAP
markers_to_plot <- c("HLA-ABC","HH3", "SMA","H3K27me3","panCK","Vimentin","CK8_18_19","CD68", "CK14","p53", "CD20", "GLUT1","CD3","p-HH3", "CD11c",
                     "Bcl-2" ,"HER2", "CK5" ,"CD274", "AR", "CD45","GATA3","CK7","CD8a", 
                     "CAIX" ,"Fibronectin", "Ki-67_Er168","EGFR","p-S6","CD4","CD31_vWF",
                     "mTOR","E/P-Cadherin","p-Rb","cleaved_CP","DNA1","Ki-67_Pt198" )

p.list <- list()
for(i in markers_to_plot){
  p.list[[i]] <- plotUMAP(sce_sub, colour_by = i, by_exprs_values = "scaled", 
                          point_size = 0.5, point_alpha = 1)  + scale_colour_gradient2(name = i, low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-3, 3), na.value = "#b2182b")
}

plot_grid(plotlist = p.list, ncol =8)
```

# Run graph-based clustering (Phenograph)

Use normalized counts for clustering.
Include only markers that are relevant for the general distinction between tumor and stromal cells

```{r phenograph-clustering}

cluster_counts_normalized <- subset(assay(sce, "normalized"), rownames(assay(sce, "normalized")) %in% markers_include)

g <- buildSNNGraph(cluster_counts_normalized, k=5, type = "jaccard")
clust<- igraph::cluster_louvain(g)$membership
table(clust)


# add cluster information to sce object
sce$phenograph_cluster_5 <- as.factor(clust)

# add cluster information to sce_sub object
sub_cellnames <- colnames(sce_sub)
cluster_sub <- sce[,sub_cellnames]$phenograph_cluster_5
sce_sub$phenograph_cluster_5 <- cluster_sub

```


## Cluster annotation

Annotate each cluster with three metadata columns:
`class` is either epithelial, stromal, immune or undefined,`celltype` is a more precise celltype annotation
Clusters high for CD20, CD4,CD3, CD45, CD8a, CD11c, CD68 expression were identified as immune cells. The CD31+vWF high cluster was used for marking endothelial cells. Clusters with high expression of SMA, vimentin, and fibronectin and low expression of panCK or E/P-cadherin were considered to be stromal cells. All clusters expressing panCK or E/P-cadherin as evaluated on the density plot were assigned as epithelial clusters, except where CK5 and CK14 markers were used for indicating basal cell clusters.


```{r cluster_annotation, fig.width=12, fig.height=12}
labels <- read.csv("data/cluster_annotations/tumorPanel_phenograph_k5_initial_cluster_annotation.csv")

label_class <- labels$class[match(sce$phenograph_cluster_5, labels$cluster)]
sce$cell_class <- label_class

label_celltype <- labels$celltype[match(sce$phenograph_cluster_5, labels$cluster)]
sce$celltype <- label_celltype


#plot on UMAP
label_class_sub <- labels$class[match(sce_sub$phenograph_cluster_5, labels$cluster)]
sce_sub$cell_class <- label_class_sub
label_celltype_sub <- labels$celltype[match(sce_sub$phenograph_cluster_5, labels$cluster)]
sce_sub$celltype <- label_celltype_sub


dittoDimPlot(sce_sub, var = "cell_class", reduction.use = "UMAP", size = 0.5)
dittoDimPlot(sce_sub, var = "celltype", reduction.use = "UMAP", size = 0.5)
dittoDimPlot(sce_sub, var = "phenograph_cluster_5", reduction.use = "UMAP", size = 0.5)
```

## Heatmap of cluster marker expression

```{r  heatmap-mean-marker-expression-per-cluster}
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% markers_include, "type", "state")
rowData(sce)$marker_class <- ifelse(rownames(sce) %in% c("DNA1", "DNA2"), "none", rowData(sce)$marker_class)

sce$sample_type[sce$Tissue.Type == 'PT'] <- "PT"
sce$sample_type[sce$Tissue.Type == 'Liver'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Bone'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Brain'] <- "M"
sce$sample_type[sce$Tissue.Type == 'SoftTissue'] <- "M"
sce$sample_type <- factor(sce$sample_type)
sce$sample_id <- factor(paste(sce$Patient.ID, sce$sample_type)) 


sce$cluster_id <- sce$phenograph_cluster_5
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))

plotMultiHeatmap(sce, 
hm1 = "type", hm2 = "state", 
k = "custom", col_dend = c(TRUE, TRUE), assay = "counts", scale = "last", row_anno= FALSE,bars = TRUE, perc = TRUE, fun = "mean")

```

# Evaluate panCK and Ecadherin expression on a density plot to make sure that epithelial clusters will have expression of these markers

```{r panCK-Ecad- marker-distribution-per-cluster}
plotClusterExprs(sce, k = "custom", features = c("panCK", "E/P-Cadherin"))
```

### Mark matched patient samples with new column in sce

```{r matched-samples-annotation}

# Subset the metadata to exclude images that have been excluded due to small cell numbers
image_metadata <- read.csv("data/tumor_images_patient_metadata_encoded.csv")
image_metadata <- image_metadata[!(duplicated(image_metadata$ROI_ID)),]
rownames(image_metadata) <- image_metadata$ROI_ID
images_incl <- unique(sce$ROI_ID)
image_metadata <- image_metadata[images_incl,]


# Separate PT and metastasis images
image_metadata_PT <- filter(image_metadata, Tissue.Type == "PT")
image_metadata_met <- filter(image_metadata, Tissue.Type != "PT")

# Count PT images for different locations
table_PT <- as.data.frame(table(image_metadata_PT$Patient.ID, image_metadata_PT$Location))

# Count metastasis images
table_met <- as.data.frame(table(image_metadata_met$Patient.ID, image_metadata_met$Tissue.Type ))

# Combine tables
table_comb <- rbind(table_PT, table_met)
table_comb_short <- pivot_wider(table_comb, names_from = Var2, values_from = Freq)
table_comb_short[is.na(table_comb_short)] = 0
colnames(table_comb_short)[1:4] <- c("PatientID", "PT_intratumoral", "PT_margin", "PT_stromal")
table_comb_short$anyMetastasis <- rowSums(table_comb_short[,5:8])

# Number of patients that have images from different metastasis sites
table_met_excl <- filter(table_met, Freq > 0)
test <- as.vector(table(table_met_excl$Var1))
length(test[test >1])

table_comb_short$anyPrimary <- rowSums(table_comb_short[,2:4])
# Remove patients that do not have any images from metastatic sites
table_comb_short <- table_comb_short %>% filter(anyPrimary != 0)
table_comb_short <- table_comb_short %>% filter(anyMetastasis != 0)

matched_samples <- rep(0,length(colnames(sce)))
#Indicate which patients have matched samples present in our dataset
matched_patients <- as.character(table_comb_short$PatientID)
index_matched_samples_multiple <- sce$Patient.ID %in% matched_patients
matched_samples[index_matched_samples_multiple] <- 1
sce$matched_samples <- matched_samples
sce$matched_samples <- factor(sce$matched_samples)

```

### Plot cell numbers per patient

```{r cell-numbers-per-patient}
sce1 <- filterSCE(sce,sce$matched_samples==1)

cellnumber <- as.data.frame(table(sce1$Patient.ID, sce1$Tissue.Type))
cellnumber$Var2 <- factor(cellnumber$Var2)
cellnumber$Var2 <- relevel(cellnumber$Var2, "PT")
cellnumber1 <- cellnumber[cellnumber$Freq!=0,]
cellnumber1$Freq_log <- log10(cellnumber1$Freq)

ggplot(cellnumber1, aes(x=Var2, y = Freq_log)) +
  geom_boxplot(outlier.size = 0.5, width = 0.2)+
 theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  labs(title="Number of all cells per patient", y = "Cell count (log10)")
```

### cell class proportion in intratumoral regions only (most metastasis samples have only intratumoral regions)
```{r cell-class-proportions-intratumoral-region}

sce_intratum <- sce[,which(sce1$Location == "intratumoral")]
tab <- as.data.frame(prop.table(table(sce_intratum$sample_id, sce_intratum$cell_class), margin = 1))
colnames(tab) <- c("sample_id", "cell_class", "proportion")
tab <- tab[complete.cases(tab),]

ROI_tissue <- as.data.frame(table(sce_intratum$sample_id, sce_intratum$Tissue.Type)) %>% filter(Freq > 0)
tab$tissue <- ROI_tissue$Var2[match(tab$sample_id, ROI_tissue$Var1)]
tab$tissue <- relevel(tab$tissue, "PT")

ggplot(tab, aes(x= tissue, y = proportion))+
  geom_boxplot()+
  geom_point()+
  facet_wrap(~cell_class, scales="fixed", ncol=6, strip.position="top")+
  theme(axis.ticks.x = element_blank(),
        axis.text.x=element_text(angle = 90, hjust=1),
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        strip.background = element_blank(),
        legend.background = element_blank())+
  ylab("Of total [%]")

```


# Save SCE object

```{r}
saveRDS(sce, "output/SCEs/tumor/tumorSCE_initial_clustered.rds")
```

