---
title: "TumorPanel_08_spatial_analysis"
author: "LauraKuett"
date: "02/05/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data

I will read in the T cell `SingleCellExperiment` object and work with the full
data.

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(imcRtools)
library(tidyverse)
library(S4Vectors)
library(ggalluvial)
library(pheatmap)
library(viridis)
library(compositions)
library(Rphenograph)
library(edgeR)
library(tidyverse)

# Read SingleCellExperiment object with subclustered T cells and subset T cells only
sce <- readRDS("~/switchdrive/Institution/BCmet/BCmet_pub/output/SCEs/Tcell/TcellSCE_subclustered.rds")

# Interaction information
neighbours <- read_csv("~/switchdrive/Institution/BCmet/BCmet_pub/data/immune_Object relationships.csv")
from_nodes <- paste0(neighbours$`First Image Number`, "_", neighbours$`First Object Number`)
to_nodes <- paste0(neighbours$`Second Image Number`, "_", neighbours$`Second Object Number`)
cell_ids <- paste0(sce$ImageNumber, "_", sce$CellNumber)

from_nodes <- match(from_nodes, cell_ids)
to_nodes <- match(to_nodes, cell_ids)

cur_pairs <- SelfHits(from = from_nodes[!is.na(from_nodes) & !is.na(to_nodes)],
                      to = to_nodes[!is.na(from_nodes) & !is.na(to_nodes)],
                      nnode = length(cell_ids))
colPair(sce, "neighbours") <- cur_pairs