---
title: "TumorPanel_03_subclustering"
author: "LauraKuett"
date: "2021-02-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script analyzes epithelial cells by clustering the cells into phenotypes using different clustering approaches and vizualises the clusters in umap.

# Load libraries and data

First, we will load the needed libraries and the epithelial `SingleCellExperiment` object, that was created in the previous section.

```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(dittoSeq)
library(scales)
library(tidyverse)
library(ggpubr)
library(cowplot)
library(scran)
library(CATALYST)
library(edgeR)
library(bluster)
```


```{r set-up-sce}
# Read SingleCellExperiment object
sce_initial  <- readRDS("output/SCEs/tumor/tumorSCE_initial_clustered.rds")

sce <- filterSCE(sce_initial, celltype == "Epithelial")


clustering_markers <- c("HLA-ABC","panCK","Vimentin","CK8_18_19",
                     "ER", "CK14","p53", "GLUT1",
                     "Bcl-2" ,"HER2", "CK5" ,"CD274", "AR","GATA3","CK7",
                     "CAIX" ,"Fibronectin", "Ki-67_Er168","EGFR","p-S6",
                     "mTOR","E/P-Cadherin","p-Rb","cleaved_CP")

#REMOVE ER - umap plotting not working for ER
markers_include_umap <- c("HLA-ABC","panCK","Vimentin","CK8_18_19",
                     "CK14","p53", "GLUT1",
                     "Bcl-2" ,"HER2", "CK5" ,"CD274", "AR","GATA3","CK7",
                     "CAIX" ,"Fibronectin", "Ki-67_Er168","EGFR","p-S6",
                     "mTOR","E/P-Cadherin","p-Rb","cleaved_CP")

#set up factors for CATALYST PACKAGE
sce$Tissue.Type <- relevel(sce$Tissue.Type, "PT")
sce$sample_type[sce$Tissue.Type == 'PT'] <- "PT"
sce$sample_type[sce$Tissue.Type == 'Liver'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Bone'] <- "M"
sce$sample_type[sce$Tissue.Type == 'Brain'] <- "M"
sce$sample_type[sce$Tissue.Type == 'SoftTissue'] <- "M"
sce$sample_type <- factor(sce$sample_type)
sce$sample_id <- factor(paste(sce$Patient.ID, sce$sample_type))
sce$Patient.ID <- factor(sce$Patient.ID)
sce$N <- factor(sce$N)
sce$T <- factor(sce$T)
sce$M <- factor(sce$M)
sce$Grade <- factor(sce$Grade)
sce$Tissue.Type <- factor(sce$Tissue.Type)
sce$PT <- factor(sce$PT)
sce$ER <- factor(sce$ER)
sce$PR <- factor(sce$PR)
sce$Her2 <- factor(sce$Her2)
sce$Site.of.metastasis <- factor(sce$Site.of.metastasis)
sce$molecular.subtype <- factor(sce$molecular.subtype)
sce$condition <- factor(sce$Tissue.Type)


# Call clustering markers type, and others state markers as expected by CATALYST 
all_markers <- rownames(rowData(sce))
clustering_cols <- match(clustering_markers,all_markers)
marker_class <- rep("state", length(rownames(rowData(sce))))
marker_class[clustering_cols] <- "type"
marker_class <- factor(marker_class, levels = c("type", "state"))
rowData(sce)$marker_class <- marker_class

```


```{r graph-based-clustering}
cluster_counts_normalized <- subset(assay(sce, "normalized"), rownames(assay(sce, "normalized")) %in% clustering_markers)

g <- buildSNNGraph(cluster_counts_normalized, k=20, type = "jaccard")
clust<- igraph::cluster_louvain(g)$membership
table(clust)

# add cluster information to sce object
sce$epithelial_phenograph_clusters <- as.factor(clust)
sce$cluster_id <- sce$epithelial_phenograph_clusters

#metadata field has to be added for each clustering type ad hoc
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))


saveRDS(sce, "output/SCEs/tumor/tumorSCE_epithelial_clustered.rds")

```


```{r heatmap-median-expression-tumor-phenotypes-graph-based, fig.width=15, fig.height=10}

plotExprHeatmap(sce1, features = "type",by = "cluster_id", k = "custom",assay = "scaled",
    scale = "last", q = 0.01, perc = TRUE,bars = TRUE, col_dend = FALSE,col_clust = TRUE,row_anno= FALSE, distance = "euclidean", linkage = "ward.D2")

```


```{r flowSOM-clustering-tumor}

#Run flowSOM on scaled asinh transformed counts. CATALYST takes exprs assay as default, thus need to change the assays. 
assay(sce, "asinh") <- assay(sce, "exprs")
assay(sce, "exprs") <-assay(sce, "scaled")
sce <- cluster(sce, features = "type", xdim = 5, ydim = 9, maxK = 20,  verbose = FALSE, seed = 999) 
sce <- cluster(sce, features = "type", xdim = 5, ydim = 9, maxK = 20,  verbose = FALSE, seed = 999) 


sce$flowSOM_clusters <- sce$cluster_id
#change assays back for downstream analysis
assay(sce, "exprs") <- assay(sce, "asinh")


sce$cluster_id <- sce$flowSOM_clusters
#metadata field has to be added for each clustering type ad hoc
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))
```

```{r heatmap-median-expression-tumor-phenotypes-flowSOM, fig.width=15, fig.height=10}
plotExprHeatmap(sce1, features = "type",by = "cluster_id", k = "custom",assay = "scaled",
    scale = "last", q = 0.01, perc = TRUE,bars = TRUE, col_dend = FALSE,col_clust = TRUE,row_anno= FALSE, distance = "euclidean", linkage = "ward.D2")


```

```{r kmeans-clustering}
cluster_counts_normalized <- subset(assay(sce, "normalized"), rownames(assay(sce, "normalized")) %in% clustering_markers)
cluster_counts_normalized <- t(cluster_counts_normalized)
cluster_counts_normalized <- round(cluster_counts_normalized,3)

set.seed(42)
Kmeans.output <- kmeans(cluster_counts_normalized,nstart=3,  centers =45, iter.max = 15 )

sce$kmeans_clusters <- as.factor(Kmeans.output$cluster)
sce$cluster_id <- sce$kmeans_clusters
#metadata field has to be added for each clustering type ad hoc
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))

```

```{r heatmap-median-expression-tumor-phenotypes-kMeans, fig.width=15, fig.height=10}
plotExprHeatmap(sce1, features = "type",by = "cluster_id", k = "custom",assay = "scaled",
    scale = "last", q = 0.01, perc = TRUE,bars = TRUE, col_dend = FALSE,col_clust = TRUE,row_anno= FALSE, distance = "euclidean", linkage = "ward.D2")
```



```{r heatmap-median-expression-tumor-phenotypes-flowSOM-final-clustering-20,fig.width=15, fig.height=10}
#Run flowSOM on scaled asinh transformed counts. CATALYST takes exprs assay as default, thus need to change the assays. 
assay(sce, "asinh") <- assay(sce, "exprs")
assay(sce, "exprs") <-assay(sce, "scaled")
sce <- cluster(sce, features = "type", xdim = 4, ydim = 5, maxK = 3,  verbose = FALSE, seed = 999) 
sce <- cluster(sce, features = "type", xdim = 4, ydim = 5, maxK = 3,  verbose = FALSE, seed = 999) 

sce$flowSOM_big_clusters <- sce$cluster_id

#change assays back for downstream analysis
assay(sce, "exprs") <- assay(sce, "asinh")

sce$cluster_id <- sce$flowSOM_big_clusters
#metadata field has to be added for each clustering type ad hoc
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))

plotExprHeatmap(sce1, features = "type",by = "cluster_id", k = "custom",assay = "scaled",
    scale = "last", q = 0.01, perc = TRUE,bars = TRUE, col_dend = FALSE,col_clust = TRUE,row_anno= FALSE, distance = "euclidean", linkage = "ward.D2")

# add flowSOM 20 clusters to the initial SCE object

sce_initial$celltypeTumorClusters <- sce_initial$celltype
sce_initial$celltypeTumorClusters[sce_initial$celltypeTumorClusters == "Epithelial"]<- sce$flowSOM_big_clusters
saveRDS(sce_initial, "output/SCEs/tumor/tumorSCE_initial_clustered_with_tumor_clusters.rds")


```


```{r save-sce}
saveRDS(sce, "output/SCEs/tumor/tumorSCE_epithelial_clustered.rds")

```


## Calculate UMAP on a subset of the cells

```{r compute-UMAP-epithelial-cells}

sce$cellID <- rownames(colData(sce))

# For images with less than 300 cells: Use all cells
as.data.frame(table(sce$ImageNumber)) %>% 
  filter(Freq < 300) %>% 
  pull(Var1) -> small.images
cell_idx_smallImages <- sce[, which(sce$ImageNumber %in% small.images)]$cellID

# For image with more than 300 cells: Randomly sample 300 cells
colData(sce[,which(!(sce$ImageNumber %in% small.images))]) %>%
  as.data.frame() %>%
  group_by(ImageNumber) %>%
  sample_n(300) %>% 
  pull(cellID) -> cell_idx_largeImages

cell_idx <- c(cell_idx_largeImages, cell_idx_smallImages)
sce_sub <- sce[,cell_idx]

# Run Dimensional reduction
sce_sub <- runUMAP(sce_sub, exprs_values = "normalized",n_neighbors =15,
                     subset_row = rownames(assay(sce_sub, "normalized")) %in% clustering_markers)

```


## Visualize clusters on UMAP


```{r plot-clusters-umap-flowSOM}

sce$cluster_id <- sce$flowSOM_clusters
# add cluster information to sce_sub object
sub_cellnames <- colnames(sce_sub)
cluster_sub <- sce[,sub_cellnames]$cluster_id
sce_sub$cluster_id <- cluster_sub

dittoDimPlot(sce_sub, var = "cluster_id", reduction.use = "UMAP", size = 0.5, do.label = TRUE, opacity = 0.5)

```

```{r plot-clusters-umap-kMeans}

sce$cluster_id <- sce$kmeans_clusters

# add cluster information to sce_sub object
sub_cellnames <- colnames(sce_sub)
cluster_sub <- sce[,sub_cellnames]$cluster_id
sce_sub$cluster_id <- cluster_sub

dittoDimPlot(sce_sub, var = "cluster_id", reduction.use = "UMAP", size = 0.5, do.label = TRUE, opacity = 0.5)
```

```{r plot-clusters-umap-fgraph-based}

sce$cluster_id <- sce$epithelial_phenograph_clusters

# add cluster information to sce_sub object
sub_cellnames <- colnames(sce_sub)
cluster_sub <- sce[,sub_cellnames]$cluster_id
sce_sub$cluster_id <- cluster_sub

dittoDimPlot(sce_sub, var = "cluster_id", reduction.use = "UMAP", size = 0.5, do.label = TRUE, opacity = 0.5)

```

```{r plot-clusters-umap-fsom20-final}

sce$cluster_id <- sce$flowSOM_big_clusters

# add cluster information to sce_sub object
sub_cellnames <- colnames(sce_sub)
cluster_sub <- sce[,sub_cellnames]$cluster_id
sce_sub$cluster_id <- cluster_sub

dittoDimPlot(sce_sub, var = "cluster_id", reduction.use = "UMAP", size = 0.5, do.label = TRUE, opacity = 0.5)

```

## Visualize features on UMAP

```{r visualize-tumor-features-umap}

dittoDimPlot(sce_sub, var = "Patient.ID", reduction.use = "UMAP", size = 0.5, opacity = 0.5)

dittoDimPlot(sce_sub, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$tissuetype, opacity = 0.3) 

dittoDimPlot(sce_sub, var = "Tissue.Type", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$tissuetype, split.by = "Tissue.Type", opacity = 0.5) 

dittoDimPlot(sce_sub, var = "Location", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$Location, opacity = 0.3)

dittoDimPlot(sce_sub, var = "Location", reduction.use = "UMAP", size = 0.5, 
                               color.panel = metadata(sce)$colour_vectors$Location, split.by = "Location",opacity = 0.5)
```

## Visualize expression on UMAP

Visualize the expression of each marker.

### 1. Arcsinh-transformed counts

```{r arcsinh-expression-tumor-markers-umap, fig.width=15, fig.height=10}
p.list <- list()
for(i in markers_include_umap){
  p.list[[i]] <- plotUMAP(sce_sub, colour_by = i, by_exprs_values = "exprs", 
                          point_size = 0.5, point_alpha = 1)
}

plot_grid(plotlist = p.list, ncol = 7)
```

