---
title: "TumorPanel_06_permutation_for_similarity_indices"
author: "LauraKuett"
date: "2021-02-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


## Introduction

This script calculates ecological diversity measures for different clustering type as in the R file TumorPanel_ecological_diversity_indices_calculations but additionally does a random permutation analysis.
For random permutation the patient labels are shuffled among primary tumors and separately for metastatic sites and then for each matched pair a distance measure is calculated.For significance testing, the median distance across the whole cohort after every permutation step is calculated to get a distribution of median distances. Then the real cohort median distance value is compared to the randomly permuted median distances.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and data
```{r read-libraries-and-data, message=FALSE, results="hide"}
# Load libraries
library(ggplot2)
library(SingleCellExperiment)
library(scater)
library(viridis)
library(RColorBrewer)
library(dittoSeq)
library(scales)
library(tidyverse)
library(ggpubr)
library(cowplot)
library(scran)
library(CATALYST)
library(edgeR)
library(compositions)
library(vegan)
library(ComplexHeatmap)

#setwd("projects/BCmet/workflowr_publication/BCmet_pub")
# Read SingleCellExperiment object
sce_epithelial <- readRDS("output/SCEs/tumor/tumorSCE_epithelial_clustered.rds")


markers_to_analyse<- c("HLA-ABC","HH3", "SMA","H3K27me3","panCK","Vimentin","CK8_18_19",
                     "ER", "CK14","p53", "GLUT1","Bcl-2" ,"HER2", "CK5" ,"CD274", "AR","GATA3","CK7", 
                     "CAIX" ,"Fibronectin", "Ki-67_Er168","EGFR","p-S6",
                     "mTOR","E/P-Cadherin","p-Rb","cleaved_CP","DNA1")

clustering_markers <- c("HLA-ABC","panCK","Vimentin","CK8_18_19",
                     "ER", "CK14","p53", "GLUT1",
                     "Bcl-2" ,"HER2", "CK5" ,"CD274", "AR","GATA3","CK7",
                     "CAIX" ,"Fibronectin", "Ki-67_Er168","EGFR","p-S6",
                     "mTOR","E/P-Cadherin","p-Rb","cleaved_CP")

sce1 <- filterSCE(sce_epithelial, rownames(sce_epithelial) %in% markers_to_analyse)
#filter only patients with matched samples
sce2 <-filterSCE(sce1, matched_samples==1)
#Check number of cells per patient sample:
counts_per_sample <- table(sce2$sample_type, sce2$Patient.ID)
samples_with_few_cells <- colSums(counts_per_sample<200)
patient_to_exclude <- rownames(data.frame(samples_with_few_cells[samples_with_few_cells==1]))

unique(sce2$Site.of.metastasis[sce2$Patient.ID %in% patient_to_exclude])

sce <- filterSCE(sce2, !(sce2$Patient.ID %in% patient_to_exclude))

#Repeat tissue type as condition for plotting functions in CATALYST
sce$Tissue.Type <- relevel(sce$Tissue.Type, "PT")
sce$condition <- factor(sce$Tissue.Type)
sce$Patient.ID <- factor(sce$Patient.ID)
sce$Location <- factor(sce$Location)

colSums(table(sce$Patient.ID,sce$Site.of.metastasis)!=0)
colSums(table(sce$Patient.ID,sce$Tissue.Type)!=0)
colSums(table(sce$Patient.ID,sce$molecular.subtype)!=0)
```


### Aitchison distance for main flowSOM-20 clusters


```{r fsom}

#big flowSOM clusters
sce$cluster_id <- sce$flowSOM_big_clusters
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))


diversity_metadata <- read.csv("data/patient_metadata_final_big_flowSOM.csv")
rownames(diversity_metadata) <- diversity_metadata$Patient.ID
diversity_metadata <- diversity_metadata[diversity_metadata$Patient.ID %in% sce$Patient.ID,]

sce_filtered <- filterSCE(sce,Tissue.Type!="PT")
fq_per_patient_met <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_met[,colSums(fq_per_patient_met, na.rm = TRUE) == 0]<- NA
abundances_per_patient_met <- round(apply(fq_per_patient_met,2,function(x){x/sum(x)}),5)
abundances_per_patient_met[,colSums(abundances_per_patient_met, na.rm = TRUE) == 0]<- NA
fq_per_patient_met[abundances_per_patient_met == 'NA']<- NA


sce_filtered <- filterSCE(sce,Tissue.Type=="PT")
fq_per_patient_pt <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_pt[,colSums(fq_per_patient_pt, na.rm = TRUE) == 0]<- NA
abundances_per_patient_PT <- round(apply(fq_per_patient_pt,2,function(x){x/sum(x)}),5)
abundances_per_patient_PT[,colSums(abundances_per_patient_PT, na.rm = TRUE) == 0]<- NA
fq_per_patient_pt[abundances_per_patient_PT == 'NA']<- NA


x_pt = t(fq_per_patient_pt)
y_met = t(fq_per_patient_met)

f <- log(prop.table(x_pt + 1, 1), 2)
x <- f-rowMeans(f)
x_log_pt <- f-rowMeans(f) #need later
f <- log(prop.table(y_met + 1, 1), 2)
y <- f-rowMeans(f)
y_log_met <- f-rowMeans(f) #need later


for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L2_transformed[diversity_metadata$Patient.ID == patientID] <-jc
 
}

x = t(abundances_per_patient_PT)
y = t(abundances_per_patient_met)

x = round(x*100,1)
x[x<1] <- 0
y = round(y*100,1)
y[y<1] <- 0

for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L1_proportions[diversity_metadata$Patient.ID == patientID] <-jc
 if(any(is.na(abundances))){
   diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-NA
 }else{
 jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
 diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-jc
 diversity_metadata$dice[diversity_metadata$Patient.ID == patientID] <-(2*jc)/(1+jc)
 }
}


diversity_metadata_2 <-diversity_metadata

### random permutation part 

diversity_metadata_random_Aitchison<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_Manhattan<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_jaccard<-diversity_metadata[,0,drop=FALSE]


for (zz in 1:1000) {
  
        diversity_metadata_random_Aitchison[,zz]<-0
        diversity_metadata_random_Manhattan[,zz]<-0
        diversity_metadata_random_jaccard[,zz]<-0
        
        x <- x_log_pt[,sample(ncol(x_log_pt))]
        colnames(x) <- colnames( x_log_pt)
        
        y <-y_log_met[,sample(ncol(y_log_met))]
        colnames(y) <- colnames(y_log_met)
        
        abundances_per_patient_PT_r <- abundances_per_patient_PT[,sample(ncol(abundances_per_patient_PT))]
        colnames(abundances_per_patient_PT_r) <- colnames(abundances_per_patient_PT)
        
        abundances_per_patient_met_r <- abundances_per_patient_met[,sample(ncol(abundances_per_patient_met))]
        colnames(abundances_per_patient_met_r) <- colnames(abundances_per_patient_met)
        
  
        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Aitchison[rownames(diversity_metadata_random_Aitchison) == patientID,zz] <-jc
         
        }
        
        x = t(abundances_per_patient_PT_r)
        y = t(abundances_per_patient_met_r)
        
        x = round(x*100,1)
        x[x<1] <- 0
        y = round(y*100,1)
        y[y<1] <- 0

        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Manhattan[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         if(any(is.na(abundances))){
           diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-NA
         }else{
         jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
         diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         }
        }
}

```


### For Aitchison distance plot distribution of real mean and permuted mean and median similairty comparison for the whole cohort

```{r Aitchison-permutation-fsom20}
x_obs <- diversity_metadata_2$L2_transformed
y_sim <- diversity_metadata_random_Aitchison


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance)) ###


real_mean <- data.frame(median(x_obs, na.rm = TRUE)) ###
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =",sum(means<=real_mean$mean)/length(means))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_histogram(aes(y=..density..), position="identity", alpha=0.5)+ 
  scale_color_grey()+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean, color=type),
             linetype="dashed")+theme(legend.position="top")


means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=20, y=1, label=sig,
              color="red")


```

### Manhattan for main flowSOM-20 clusters

```{r bray-curtis-permutation-fsome20}

x_obs <- diversity_metadata_2$L1_proportions
y_sim <- diversity_metadata_random_Manhattan


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")
```

### Jaccard for main flowSOM-20 clusters

```{r jaccard-permutation-fsom20}

x_obs <- diversity_metadata_2$jaccard
y_sim <- diversity_metadata_random_jaccard


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance)) ###


real_mean <- data.frame(median(x_obs, na.rm = TRUE))###
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")
```




### Same for other clustering types

# k means

```{r kmeans_clusters}

#big flowSOM clusters
sce$cluster_id <- sce$kmeans_clusters
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))


diversity_metadata <- read.csv("data/patient_metadata_final_big_flowSOM.csv")
rownames(diversity_metadata) <- diversity_metadata$Patient.ID
diversity_metadata <- diversity_metadata[diversity_metadata$Patient.ID %in% sce$Patient.ID,]

sce_filtered <- filterSCE(sce,Tissue.Type!="PT")
fq_per_patient_met <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_met[,colSums(fq_per_patient_met, na.rm = TRUE) == 0]<- NA
abundances_per_patient_met <- round(apply(fq_per_patient_met,2,function(x){x/sum(x)}),5)
abundances_per_patient_met[,colSums(abundances_per_patient_met, na.rm = TRUE) == 0]<- NA
fq_per_patient_met[abundances_per_patient_met == 'NA']<- NA


sce_filtered <- filterSCE(sce,Tissue.Type=="PT")
fq_per_patient_pt <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_pt[,colSums(fq_per_patient_pt, na.rm = TRUE) == 0]<- NA
abundances_per_patient_PT <- round(apply(fq_per_patient_pt,2,function(x){x/sum(x)}),5)
abundances_per_patient_PT[,colSums(abundances_per_patient_PT, na.rm = TRUE) == 0]<- NA
fq_per_patient_pt[abundances_per_patient_PT == 'NA']<- NA


x_pt = t(fq_per_patient_pt)
y_met = t(fq_per_patient_met)

f <- log(prop.table(x_pt + 1, 1), 2)
x <- f-rowMeans(f)
x_log_pt <- f-rowMeans(f) #need later
f <- log(prop.table(y_met + 1, 1), 2)
y <- f-rowMeans(f)
y_log_met <- f-rowMeans(f) #need later


for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L2_transformed[diversity_metadata$Patient.ID == patientID] <-jc
 
}

x = t(abundances_per_patient_PT)
y = t(abundances_per_patient_met)
x = round(x*100,1)
x[x<1] <- 0
y = round(y*100,1)
y[y<1] <- 0

for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L1_proportions[diversity_metadata$Patient.ID == patientID] <-jc
 if(any(is.na(abundances))){
   diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-NA
 }else{
 jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
 diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-jc
 diversity_metadata$dice[diversity_metadata$Patient.ID == patientID] <-(2*jc)/(1+jc)
 }
}


diversity_metadata_2 <-diversity_metadata

### random permutation part 

diversity_metadata_random_Aitchison<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_Manhattan<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_jaccard<-diversity_metadata[,0,drop=FALSE]


for (zz in 1:10000) {
  
        diversity_metadata_random_Aitchison[,zz]<-0
        diversity_metadata_random_Manhattan[,zz]<-0
        diversity_metadata_random_jaccard[,zz]<-0
        
        x <- x_log_pt[,sample(ncol(x_log_pt))]
        colnames(x) <- colnames( x_log_pt)
        
        y <-y_log_met[,sample(ncol(y_log_met))]
        colnames(y) <- colnames(y_log_met)
        
        abundances_per_patient_PT_r <- abundances_per_patient_PT[,sample(ncol(abundances_per_patient_PT))]
        colnames(abundances_per_patient_PT_r) <- colnames(abundances_per_patient_PT)
        
        abundances_per_patient_met_r <- abundances_per_patient_met[,sample(ncol(abundances_per_patient_met))]
        colnames(abundances_per_patient_met_r) <- colnames(abundances_per_patient_met)
        
  
        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Aitchison[rownames(diversity_metadata_random_Aitchison) == patientID,zz] <-jc
         
        }
        
        x = t(abundances_per_patient_PT_r)
        y = t(abundances_per_patient_met_r)
        x = round(x*100,1)
        x[x<1] <- 0
        y = round(y*100,1)
        y[y<1] <- 0
        
        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Manhattan[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         if(any(is.na(abundances))){
           diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-NA
         }else{
         jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
         diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         }
        }
}

```


###  Aitchison distance

```{r Aitchison-permutation-km}

x_obs <- diversity_metadata_2$L2_transformed
y_sim <- diversity_metadata_random_Aitchison


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=70, y=1, label=sig,
              color="red")


```

### Manhattan

```{r bray-curtis-permutation-km}

x_obs <- diversity_metadata_2$L1_proportions
y_sim <- diversity_metadata_random_Manhattan


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")


```

### Jaccard

```{r jaccard-permutation-km}

x_obs <- diversity_metadata_2$jaccard
y_sim <- diversity_metadata_random_jaccard


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")

```



# grpah based clustering



```{r graph-based}

#big flowSOM clusters
sce$cluster_id <- sce$epithelial_phenograph_clusters
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))


diversity_metadata <- read.csv("data/patient_metadata_final_big_flowSOM.csv")
rownames(diversity_metadata) <- diversity_metadata$Patient.ID
diversity_metadata <- diversity_metadata[diversity_metadata$Patient.ID %in% sce$Patient.ID,]

sce_filtered <- filterSCE(sce,Tissue.Type!="PT")
fq_per_patient_met <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_met[,colSums(fq_per_patient_met, na.rm = TRUE) == 0]<- NA
abundances_per_patient_met <- round(apply(fq_per_patient_met,2,function(x){x/sum(x)}),5)
abundances_per_patient_met[,colSums(abundances_per_patient_met, na.rm = TRUE) == 0]<- NA
fq_per_patient_met[abundances_per_patient_met == 'NA']<- NA


sce_filtered <- filterSCE(sce,Tissue.Type=="PT")
fq_per_patient_pt <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_pt[,colSums(fq_per_patient_pt, na.rm = TRUE) == 0]<- NA
abundances_per_patient_PT <- round(apply(fq_per_patient_pt,2,function(x){x/sum(x)}),5)
abundances_per_patient_PT[,colSums(abundances_per_patient_PT, na.rm = TRUE) == 0]<- NA
fq_per_patient_pt[abundances_per_patient_PT == 'NA']<- NA


x_pt = t(fq_per_patient_pt)
y_met = t(fq_per_patient_met)

f <- log(prop.table(x_pt + 1, 1), 2)
x <- f-rowMeans(f)
x_log_pt <- f-rowMeans(f) #need later
f <- log(prop.table(y_met + 1, 1), 2)
y <- f-rowMeans(f)
y_log_met <- f-rowMeans(f) #need later


for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L2_transformed[diversity_metadata$Patient.ID == patientID] <-jc
 
}

x = t(abundances_per_patient_PT)
y = t(abundances_per_patient_met)

x = round(x*100,1)
x[x<1] <- 0
y = round(y*100,1)
y[y<1] <- 0

for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L1_proportions[diversity_metadata$Patient.ID == patientID] <-jc
 if(any(is.na(abundances))){
   diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-NA
 }else{
 jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
 diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-jc
 diversity_metadata$dice[diversity_metadata$Patient.ID == patientID] <-(2*jc)/(1+jc)
 }
}


diversity_metadata_2 <-diversity_metadata

### random permutation part 

diversity_metadata_random_Aitchison<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_Manhattan<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_jaccard<-diversity_metadata[,0,drop=FALSE]


for (zz in 1:10000) {
  
        diversity_metadata_random_Aitchison[,zz]<-0
        diversity_metadata_random_Manhattan[,zz]<-0
        diversity_metadata_random_jaccard[,zz]<-0
        
        x <- x_log_pt[,sample(ncol(x_log_pt))]
        colnames(x) <- colnames( x_log_pt)
        
        y <-y_log_met[,sample(ncol(y_log_met))]
        colnames(y) <- colnames(y_log_met)
        
        abundances_per_patient_PT_r <- abundances_per_patient_PT[,sample(ncol(abundances_per_patient_PT))]
        colnames(abundances_per_patient_PT_r) <- colnames(abundances_per_patient_PT)
        
        abundances_per_patient_met_r <- abundances_per_patient_met[,sample(ncol(abundances_per_patient_met))]
        colnames(abundances_per_patient_met_r) <- colnames(abundances_per_patient_met)
        
  
        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Aitchison[rownames(diversity_metadata_random_Aitchison) == patientID,zz] <-jc
         
        }
        
        x = t(abundances_per_patient_PT_r)
        y = t(abundances_per_patient_met_r)
        
        x = round(x*100,1)
        x[x<1] <- 0
        y = round(y*100,1)
        y[y<1] <- 0
        
        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Manhattan[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         if(any(is.na(abundances))){
           diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-NA
         }else{
         jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
         diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         }
        }
}

```


### Aitchison distance

```{r Aitchison-permutation-gb}

x_obs <- diversity_metadata_2$L2_transformed
y_sim <- diversity_metadata_random_Aitchison


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=70, y=1, label=sig,
              color="red")

```

### Manhattan

```{r bray-curtis-permutation-gb}

x_obs <- diversity_metadata_2$L1_proportions
y_sim <- diversity_metadata_random_Manhattan


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")


```

### Jaccard

```{r jaccard-permutation-gb}

x_obs <- diversity_metadata_2$jaccard
y_sim <- diversity_metadata_random_jaccard


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")

```



# fsom-45


```{r fsom-45}

#big flowSOM clusters
sce$cluster_id <- sce$flowSOM_clusters
metadata(sce)$cluster_codes <- data.frame(custom = factor(levels(sce$cluster_id)), levels = levels(sce$cluster_id))


diversity_metadata <- read.csv("data/patient_metadata_final_big_flowSOM.csv")
rownames(diversity_metadata) <- diversity_metadata$Patient.ID
diversity_metadata <- diversity_metadata[diversity_metadata$Patient.ID %in% sce$Patient.ID,]

sce_filtered <- filterSCE(sce,Tissue.Type!="PT")
fq_per_patient_met <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_met[,colSums(fq_per_patient_met, na.rm = TRUE) == 0]<- NA
abundances_per_patient_met <- round(apply(fq_per_patient_met,2,function(x){x/sum(x)}),5)
abundances_per_patient_met[,colSums(abundances_per_patient_met, na.rm = TRUE) == 0]<- NA
fq_per_patient_met[abundances_per_patient_met == 'NA']<- NA


sce_filtered <- filterSCE(sce,Tissue.Type=="PT")
fq_per_patient_pt <- table(sce_filtered$cluster_id, sce_filtered$Patient.ID)
fq_per_patient_pt[,colSums(fq_per_patient_pt, na.rm = TRUE) == 0]<- NA
abundances_per_patient_PT <- round(apply(fq_per_patient_pt,2,function(x){x/sum(x)}),5)
abundances_per_patient_PT[,colSums(abundances_per_patient_PT, na.rm = TRUE) == 0]<- NA
fq_per_patient_pt[abundances_per_patient_PT == 'NA']<- NA


x_pt = t(fq_per_patient_pt)
y_met = t(fq_per_patient_met)

f <- log(prop.table(x_pt + 1, 1), 2)
x <- f-rowMeans(f)
x_log_pt <- f-rowMeans(f) #need later
f <- log(prop.table(y_met + 1, 1), 2)
y <- f-rowMeans(f)
y_log_met <- f-rowMeans(f) #need later


for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L2_transformed[diversity_metadata$Patient.ID == patientID] <-jc
 
}

x = t(abundances_per_patient_PT)
y = t(abundances_per_patient_met)

x = round(x*100,1)
x[x<1] <- 0
y = round(y*100,1)
y[y<1] <- 0

for(i in 1:nrow(x)){
 patientID <-rownames(x)[i]
 abundances <- rbind( x[patientID,], y[patientID,])
 jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
 diversity_metadata$L1_proportions[diversity_metadata$Patient.ID == patientID] <-jc
 if(any(is.na(abundances))){
   diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-NA
 }else{
 jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
 diversity_metadata$jaccard[diversity_metadata$Patient.ID == patientID] <-jc
 diversity_metadata$dice[diversity_metadata$Patient.ID == patientID] <-(2*jc)/(1+jc)
 }
}


diversity_metadata_2 <-diversity_metadata

### random permutation part 

diversity_metadata_random_Aitchison<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_Manhattan<-diversity_metadata[,0,drop=FALSE]
diversity_metadata_random_jaccard<-diversity_metadata[,0,drop=FALSE]


for (zz in 1:10000) {
  
        diversity_metadata_random_Aitchison[,zz]<-0
        diversity_metadata_random_Manhattan[,zz]<-0
        diversity_metadata_random_jaccard[,zz]<-0
        
        x <- x_log_pt[,sample(ncol(x_log_pt))]
        colnames(x) <- colnames( x_log_pt)
        
        y <-y_log_met[,sample(ncol(y_log_met))]
        colnames(y) <- colnames(y_log_met)
        
        abundances_per_patient_PT_r <- abundances_per_patient_PT[,sample(ncol(abundances_per_patient_PT))]
        colnames(abundances_per_patient_PT_r) <- colnames(abundances_per_patient_PT)
        
        abundances_per_patient_met_r <- abundances_per_patient_met[,sample(ncol(abundances_per_patient_met))]
        colnames(abundances_per_patient_met_r) <- colnames(abundances_per_patient_met)
        
  
        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'euclidean', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Aitchison[rownames(diversity_metadata_random_Aitchison) == patientID,zz] <-jc
         
        }
        
        x = t(abundances_per_patient_PT_r)
        y = t(abundances_per_patient_met_r)
        
        x = round(x*100,1)
        x[x<1] <- 0
        y = round(y*100,1)
        y[y<1] <- 0
        
        for(i in 1:nrow(x)){
         patientID <-rownames(x)[i]
         abundances <- rbind( x[patientID,], y[patientID,])
         jc <- vegdist(abundances,method = 'manhattan', binary = FALSE,na.rm = TRUE)
         diversity_metadata_random_Manhattan[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         if(any(is.na(abundances))){
           diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-NA
         }else{
         jc <- vegdist(abundances,method = 'jaccard', binary = TRUE,na.rm = TRUE)
         diversity_metadata_random_jaccard[rownames(diversity_metadata_random_Aitchison),zz] <-jc
         }
        }
}

```


### Aitchsison distance

```{r Aitchison-permutation-fs45}

x_obs <- diversity_metadata_2$L2_transformed
y_sim <- diversity_metadata_random_Aitchison


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=70, y=1, label=sig,
              color="red")


```

### Manhattan

```{r bray-curtis-permutation-fs45}

x_obs <- diversity_metadata_2$L1_proportions
y_sim <- diversity_metadata_random_Manhattan


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")

```

### Jaccard

```{r jaccard-permutation-fs45}

x_obs <- diversity_metadata_2$jaccard
y_sim <- diversity_metadata_random_jaccard


df_tmp <- reshape2::melt(cbind(y_sim,x_obs))
df_tmp$type <- "simulation"

df_tmp$type[df_tmp$variable == "x_obs"] <- "observed"
colnames(df_tmp) <- c("variable", "distance", "type")
mu <- plyr::ddply(df_tmp, "type", summarise, grp.mean=median(distance))


real_mean <- data.frame(median(x_obs, na.rm = TRUE))
colnames(real_mean)<- "mean"
means <- colMedians(as.matrix(y_sim))
sig <-paste("p =", round(sum(means<=real_mean$mean)/length(means),3))


ggplot(df_tmp,aes(x=distance, color=type,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")

means <- as.data.frame(means)
colnames(means)<-"median_similarity"
ggplot(means,aes(x=median_similarity,..scaled..)) + geom_density()+ scale_color_manual(values=c("#E69F00", "#56B4E9"))+theme_classic()+
  geom_vline(data=mu, aes(xintercept=grp.mean[1], color=type[1]),
             linetype="dashed")+theme(legend.position="top")+annotate(geom="text", x=0.7, y=1, label=sig,
              color="red")
```


